---
title: "Capitulo 1 - Single-cell RNA-seq analysis"
subtitle: "Runnable R Markdown blocks for QC, clustering, bibliography annotation, curation, and h5ad export"
author: "SingleCell Pipeline"
date: "2026-07-14"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  error = FALSE
)
```

This R Markdown version is split into executable blocks so each section can be run and checked independently. It follows `workflow/capitulo1_single_cell.R` as of 2026-07-14.

Reference-transfer annotation is not run here. The active annotation path is bibliography-based (`biblio_marks.txt` -> `pbmc_harmony$celltype`). The older reference-transfer workflow is preserved separately in `workflow/capitulo1_single_cell_reference_backup.R`.

Run inside the Docker container with `/workspace` mounted to the project directory. Execute sections in order. Section 11 is interactive and has `eval=FALSE` to prevent accidental execution during a full knit.


## Configuration


Edit this block before running the pipeline.


```{r configuration}
# CONFIGURATION — Edit only this block before running the pipeline
# =============================================================================

# Directory containing the pipeline helper scripts.
# Inside the Docker container this is typically /workspace/SingleCell
PIPELINE_DIR <- "/workspace/SingleCell/workflow"

# Root directory for your project data and results.
# All result files will be written to DATA_DIR/resultados/<step>/
DATA_DIR   <- "/workspace/."
base_dir   <- file.path(DATA_DIR, "resultados")

# =============================================================================
```


## Initialization


Load helper scripts, set reproducibility options, and create output directories.


```{r initialization}
# INITIALIZATION
# =============================================================================

# ── Load helper scripts ────────────────────────────────────────────────────────
# Each file is fully documented at https://github.com/cliford2001/SingleCell
source(file.path(PIPELINE_DIR, "load_libraries.R"))          # all R packages
source(file.path(PIPELINE_DIR, "ScRNA_Analysis_Functions.R"))# analysis functions

set.seed(1807)
options(Seurat.allow.s4 = FALSE)  # required for Seurat 5 compatibility
setwd(DATA_DIR)

# ── Create per-step output directories ────────────────────────────────────────
list2env(create_pipeline_dirs(base_dir), envir = .GlobalEnv)  # creates output folders and loads their paths as variables

# output_dir is the global variable used by save_pdf / save_qc / save_vln helpers.
# It is reassigned at the start of each section to the appropriate step directory.
output_dir <- base_dir


# =============================================================================
# ████████████████████████  PART 1 — SINGLE-CELL ANALYSIS  ████████████████████
# =============================================================================
```


## Section 1 - DATA LOADING AND PRE-FILTER QC


```{r section_01_data_loading_and_pre_filter_qc}
# SECTION 1 — DATA LOADING AND PRE-FILTER QC
# =============================================================================
# Each sample is loaded from its input file and mitochondrial / chloroplast
# read fractions are computed per cell to guide filtering thresholds in
# Section 2. Pre-filter violin plots are saved to 01_qc/.
#
# ┌─ CHANGE FOR YOUR ORGANISM ──────────────────────────────────────────────────
#   Arabidopsis : mt_pattern = "^ATMG"  |  cp_pattern = "^ATCG"
#   Human       : mt_pattern = "^MT-"   |  cp_pattern = NULL
#   Mouse       : mt_pattern = "^mt-"   |  cp_pattern = NULL
# └─────────────────────────────────────────────────────────────────────────────
output_dir <- dir_01

# ── Sample manifest (CellRanger filtered_feature_bc_matrix) ──────────────────
# Add one entry per sample. Each entry needs:
#   file      — path to the filtered_feature_bc_matrix/ directory (relative to DATA_DIR)
#   label     — unique name for this sample (appears in all plots)
#   condition — experimental group this sample belongs to
samples <- list(
  list(file = "cellranger_v2/ScWT/outs/filtered_feature_bc_matrix",  label = "WT",  condition = "WT"),
  list(file = "cellranger_v2/Scpifq/outs/filtered_feature_bc_matrix", label = "pifq", condition = "pifq"))


# ── Plot colors (one color per sample label) ───────────────────────────────────
colors <- c(
  "WT"  = "#66c2a5",
  "pifq"  = "#fc8d62"
)


mt_pattern <- "^ATMG"  # Arabidopsis mitochondrial genes
cp_pattern <- "^ATCG"  # Arabidopsis chloroplast genes


# Load all samples using the helper function
seurat_list_raw <- load_seurat_samples(samples = samples,
                                       DATA_DIR = DATA_DIR,
                                       mt_pattern = mt_pattern,
                                       cp_pattern = cp_pattern)

plot_qc_batch(seurat_list_raw, colors, "qc_prefilter.pdf")


# =============================================================================

message("\n✓ SECTION 1 COMPLETE: QC pre-filter plots saved")
```


## Section 2 - CELL FILTERING AND DOUBLET DETECTION


```{r section_02_cell_filtering_and_doublet_detection}
# SECTION 2 — CELL FILTERING AND DOUBLET DETECTION
# =============================================================================
# Thresholds are set based on the pre-filter violin plots (01_qc/qc_prefilter.pdf).
# Cells below min_features or above max_mt are removed. DoubletFinder is
# applied per sample to discard putative doublets.
#
# ┌─ ADJUST THRESHOLDS AFTER INSPECTING 01_qc/qc_prefilter.pdf ─────────────────
#   min_features : minimum number of detected genes per cell (default 200)
#   max_mt       : maximum mitochondrial read percentage  (default 5 %)
# └─────────────────────────────────────────────────────────────────────────────
output_dir <- dir_01   # both pre- and post-filter QC plots go to 01_qc/

seurat_list <- filter_seurat_samples(seurat_list_raw, min_features = 200, max_mt = 5)

plot_qc_batch(seurat_list, colors, "qc_postfilter.pdf")

# Checkpoint — restore with: seurat_list <- readRDS("resultados/objects/seurat_list_postfilter.rds")
saveRDS(seurat_list, file.path(dir_objects, "seurat_list_postfilter.rds"))


# =============================================================================

message("\n✓ SECTION 2 COMPLETE: Filtering and doublet detection complete")
```


## Section 3 - MERGE AND INITIAL PREPROCESSING


```{r section_03_merge_and_initial_preprocessing}
# SECTION 3 — MERGE AND INITIAL PREPROCESSING
# =============================================================================
# Filtered samples are merged and preprocessed: log-normalization, variable
# feature selection (VST, 2,000 features), scaling, PCA (30 PCs), and UMAP.
# The resulting UMAP shows batch effects before integration.

output_dir <- dir_01

pbmc_harmony <- reduce(seurat_list, merge) %>%  # merge all samples into one object
  NormalizeData(verbose = FALSE) %>%
  FindVariableFeatures(selection.method = "vst", nfeatures = 2000, verbose = FALSE) %>%
  ScaleData(verbose = FALSE) %>%
  RunPCA(npcs = 30, verbose = FALSE) %>%
  RunUMAP(reduction = "pca", dims = 1:30, verbose = FALSE)

save_pdf(DimPlot(pbmc_harmony, group.by = "orig.ident", cols = colors),
         "umap_preharmony.pdf")

# Checkpoint — restore with: pbmc_harmony <- readRDS("resultados/objects/pbmc_harmony_preharmony.rds")
saveRDS(pbmc_harmony, file.path(dir_objects, "pbmc_harmony_preharmony.rds"))


# =============================================================================

message("\n✓ SECTION 3 COMPLETE: Merge and preprocessing complete")
```


## Section 4 - HARMONY BATCH CORRECTION


```{r section_04_harmony_batch_correction}
# SECTION 4 — HARMONY BATCH CORRECTION
# =============================================================================
# Harmony adjusts cell embeddings to remove sample-level batch effects while
# preserving biological variation. All downstream steps use the "harmony"
# reduction instead of "pca".

output_dir <- dir_01

pbmc_harmony <- pbmc_harmony %>%
  RunHarmony("orig.ident", plot_convergence = FALSE) %>%
  RunUMAP(reduction = "harmony", dims = 1:30, verbose = FALSE)

save_pdf(DimPlot(pbmc_harmony, group.by = "orig.ident", cols = colors),
         "umap_postharmony.pdf")

# Checkpoint — restore with: pbmc_harmony <- readRDS("resultados/objects/pbmc_harmony_postharmony.rds")
saveRDS(pbmc_harmony, file.path(dir_objects, "pbmc_harmony_postharmony.rds"))


# =============================================================================

message("\n✓ SECTION 4 COMPLETE: Harmony integration complete")
```


## Section 5 - RESOLUTION OPTIMIZATION


```{r section_05_resolution_optimization}
# SECTION 5 — RESOLUTION OPTIMIZATION
# =============================================================================
# Clustree tracks how Leiden communities split or merge across candidate
# resolutions. Choose the lowest resolution where clusters stabilize.
#
# ┌─ PARAMETERS ────────────────────────────────────────────────────────────────
#   resolutions_test : Leiden resolutions swept by clustree
#   → Inspect elbow_plot.pdf and clustree2.pdf before setting cluster_resolution in Section 6.
# └─────────────────────────────────────────────────────────────────────────────
resolutions_test <- c(0.15, 0.35, 0.45, 0.55, 1.0)

output_dir <- dir_02

# Elbow plot: k-means within-cluster sum of squares across candidate k values.
# This is only a diagnostic plot; it does not change the clustering parameters.
k_range <- 1:31
pca_data <- Embeddings(pbmc_harmony, "pca")[, 1:30]
wss <- sapply(
  k_range,
  function(k) kmeans(pca_data, centers = k, nstart = 4)$tot.withinss
)

elbow_plot <- ggplot(data.frame(k = k_range, wss = wss), aes(k, wss)) +
  geom_line() +
  geom_point() +
  labs(
    x = "Number of clusters (k)",
    y = "Within-cluster sum of squares"
  ) +
  theme_minimal()

save_pdf(elbow_plot, "elbow_plot.pdf", w = 18, h = 18)

clu <- pbmc_harmony %>%
  RunUMAP(reduction = "harmony", dims = 1:30, verbose = FALSE) %>%
  FindNeighbors(reduction = "harmony", dims = 1:30,
                k.param = 20, verbose = FALSE)

for (res in resolutions_test)
  clu <- FindClusters(clu, resolution = res, algorithm = 4, verbose = FALSE)

save_pdf(clustree(clu, prefix = "RNA_snn_res."), "clustree2.pdf", w = 18, h = 18)


# =============================================================================

message("\n✓ SECTION 5 COMPLETE: Elbow plot and clustree saved — inspect elbow_plot.pdf and clustree2.pdf before setting resolution")
```


## Section 6 - FINAL CLUSTERING


```{r section_06_final_clustering}
# SECTION 6 — FINAL CLUSTERING
# =============================================================================
# Apply the selected resolution for the final cluster assignment.
# After clustering, a UMAP coloured by Seurat cluster and a bar chart of
# cells per sample are saved to 02_clustering/.
#
# ┌─ SET RESOLUTION AFTER INSPECTING clustree.pdf ─────────────────────────────
#   cluster_resolution : Leiden resolution for final clustering (default 0.3)
# └─────────────────────────────────────────────────────────────────────────────
cluster_resolution <- 0.35
output_dir <- dir_02

# clu (Section 5) was temporary — re-run on pbmc_harmony to embed the final clusters
pbmc_harmony <- pbmc_harmony %>%
  RunUMAP(reduction = "harmony", dims = 1:30, verbose = FALSE) %>%
  FindNeighbors(reduction = "harmony", dims = 1:30,
                k.param = 20, verbose = FALSE) %>%
  FindClusters(resolution = cluster_resolution, algorithm = 4, verbose = FALSE)

Idents(pbmc_harmony) <- "seurat_clusters"
save_pdf(DimPlot(pbmc_harmony, group.by = "seurat_clusters", label = TRUE),
         "umap_seuratclusters.pdf")


# =============================================================================

message("\n✓ SECTION 6 COMPLETE: Final clustering complete")
```


## Section 7 - CELL-TYPE ANNOTATION


```{r section_07_cell_type_annotation}
# SECTION 7 — CELL-TYPE ANNOTATION
# =============================================================================
# Cell-type labels are assigned via bibliography-based markers
# (biblio_marks.txt): the marker table read above is crossed with
# cluster-level differential genes to assign initial labels automatically.
# Result stored in pbmc_harmony$celltype.
#
# A second strategy — reference transfer from a published Seurat object
# (e.g. an Arabidopsis leaf atlas) via FindTransferAnchors/TransferData — is
# not run by default. The full block, plus the downstream sections wired to
# it, is preserved in capitulo1_single_cell_reference_backup.R.

output_dir <- dir_03

biblio_marks_file <- file.path(DATA_DIR, "biblio_marks_custom.txt")
marker_table      <- read.table(biblio_marks_file, header = TRUE, sep = "\t", quote = "")

# ── Bibliography-based annotation ─────────────────────────────────────────────
markers <- find_markers(pbmc_harmony,
                        output_file = file.path(output_dir, "FindAllMarkers.tsv"))

pbmc_harmony <- annotate_by_markers(pbmc_harmony, markers,
                                    reference_file = biblio_marks_file)
# Annotation stored in: pbmc_harmony$celltype

plot_marker_dotplot(
  pbmc_harmony,
  marker_table,
  annot_col = "celltype", # uses the newly assigned annotation column
  outfile   = file.path(output_dir, "dotplot_marker_table_annotation_biblio.pdf"),
  width = 18, height = 18
)

save_pdf(DimPlot(pbmc_harmony, group.by = "celltype",
                 label = TRUE, repel = TRUE, raster = FALSE),
         "umap_annotation_biblio.pdf")


# =============================================================================

message("\n✓ SECTION 7 COMPLETE: Cell-type annotation complete")
```


## Section 8 - ANNOTATED CLUSTREE


```{r section_08_annotated_clustree}
# SECTION 8 — ANNOTATED CLUSTREE
# =============================================================================
# Reuses the resolution sweep object from Section 5 and overlays the final
# bibliography-based annotation on each node.

output_dir <- dir_03

# Mode: returns the most frequent annotation in each clustree node.
Mode <- function(x) {
  x <- as.character(x)
  x <- x[!is.na(x) & nzchar(x)]
  if (length(x) == 0) return(NA_character_)
  names(sort(table(x), decreasing = TRUE))[1]
}

stopifnot(exists("clu"))
stopifnot("celltype" %in% colnames(pbmc_harmony@meta.data))

celltype_label <- as.character(pbmc_harmony$celltype)
names(celltype_label) <- Cells(pbmc_harmony)
clu$celltype_label <- celltype_label[Cells(clu)]

print(table(clu$celltype_label, useNA = "ifany"))

save_pdf(
  clustree(clu, prefix = "RNA_snn_res.",
           node_label = "celltype_label", node_label_aggr = "Mode"),
  "clustree_annotated.pdf", w = 14, h = 14
)


# =============================================================================

message("\n✓ SECTION 8 COMPLETE: Annotated clustree saved")
```


## Section 9 - GENE EXPRESSION VISUALIZATION


```{r section_09_gene_expression_visualization}
# SECTION 9 — GENE EXPRESSION VISUALIZATION
# =============================================================================
# Simple inspection for one gene: violin plot across all annotated cell types
# and UMAP expression overlay.
#
# ┌─ SET GENE TO INSPECT ──────────────────────────────────────────────────────
#   gene : single gene to inspect
# └─────────────────────────────────────────────────────────────────────────────
gene <- "AT5G26000"

output_dir <- dir_04

# JoinLayers is required in Seurat 5 after merging samples.
pbmc_harmony <- JoinLayers(pbmc_harmony)
Idents(pbmc_harmony) <- "celltype"

save_vln(
  VlnPlot(pbmc_harmony, features = gene),
  "vln_gene_all.pdf"
)

save_pdf(
  FeaturePlot(pbmc_harmony, features = gene),
  "feature_gene_all.pdf"
)

# =============================================================================

message("\n✓ SECTION 9 COMPLETE: Gene expression visualization saved")
```


## Section 10 - CELL-TYPE GROUPING  [OPTIONAL]


```{r section_10_cell_type_grouping_optional}
# SECTION 10 — CELL-TYPE GROUPING  [OPTIONAL]
# =============================================================================
# Fine-grained labels are collapsed into broader categories for downstream
# analyses. Cell types NOT listed in 'grouping' keep their original label.
# Skip this section if you do not need to merge cell types.
#
# ┌─ EDIT THIS MAP TO MATCH YOUR CELL TYPES ───────────────────────────────────
#   Left side  : original label (must match exactly)
#   Right side : new broader label to assign
# └─────────────────────────────────────────────────────────────────────────────
# Leave empty if you do not want to merge labels.
#
# Example: uncomment and edit this if Section 7 produced several labels that
# should be treated as one broader group downstream.
#
# grouping <- c(
#   "Epidermis Hypocotyl.1" = "Epidermis Hypocotyl",
#   "Epidermis Hypocotyl.2" = "Epidermis Hypocotyl",
#   "Epidermis Hypocotyl.3" = "Epidermis Hypocotyl",
#   "ENDO1" = "Endodermis",
#   "ENDO2" = "Endodermis",
#   "PHLOEM AND SIEVE ELEMENT" = "PHLOEM"
# )
#
# Current default: no grouping; celltype_grouped is a straight pass-through
# of celltype.
grouping <- c()

output_dir <- dir_05

# !!! unpacks the grouping vector as named arguments to recode()
# recode() errors on an empty replacement list, so an empty `grouping` (no
# merging needed) falls back to a straight pass-through instead.
if (length(grouping) > 0) {
  pbmc_harmony$celltype_grouped <- recode(pbmc_harmony$celltype, !!!grouping)
} else {
  pbmc_harmony$celltype_grouped <- pbmc_harmony$celltype
}

save_pdf(
  DimPlot(pbmc_harmony, group.by = "celltype_grouped",  # grouped broad cell types
          label = TRUE, repel = TRUE, raster = FALSE),
  "umap_grouped.pdf"
)


# =============================================================================

message("\n✓ SECTION 10 COMPLETE: Cell-type grouping complete")
```


## Section 11 - INTERACTIVE CELL-TYPE CURATION  [OPTIONAL]


This section is intentionally interactive. Run it chunk by chunk after inspecting the previous outputs.


```{r section_11_interactive_cell_type_curation_optional, eval=FALSE}
# SECTION 11 — SUBCLUSTER MARKER INSPECTION  [OPTIONAL]
# =============================================================================
# !!! WARNING: Run this section interactively, step by step.
# !!! Do NOT source the entire script with this section active.
# Skip this section if you are satisfied with the annotation from Section 7.
#
# Purpose: pick one or more annotated cell types, subcluster them, and inspect
# marker expression across the subclusters before deciding whether any manual
# reassignment is needed. This section does NOT modify pbmc_harmony by default.

output_dir   <- dir_05
curation_col <- "celltype"
Idents(pbmc_harmony) <- curation_col

print(table(pbmc_harmony[[curation_col]], useNA = "ifany"))

# Edit this vector using exact names from the table above.
# Example options:
#   types_to_inspect <- c("Epidermis Hypocotyl.1")
#   types_to_inspect <- c("Epidermis Hypocotyl.1", "Epidermis Hypocotyl.2")
#   types_to_inspect <- names(table(pbmc_harmony[[curation_col]]))  # all labels; can be slow

types_to_inspect <- c("2","4","5","8")

subcluster_resolution <- 0.3
subcluster_dims       <- 1:20

subcluster_list    <- list()
inspection_entries <- list()

for (ct in types_to_inspect) {
  safe_name <- make.names(ct)
  file_tag  <- gsub("[^A-Za-z0-9]+", "_", ct)

  message("\nInspecting: ", ct)

  sub_obj <- subcluster_cell_type(
    pbmc_harmony,
    tipo       = ct,
    annot_col  = curation_col,
    resolution = subcluster_resolution,
    dims       = subcluster_dims
  )

  subcluster_list[[safe_name]] <- sub_obj
  print(table(sub_obj$cluster_subtipo, useNA = "ifany"))

  p_dim <- plot_subcluster_umap(sub_obj, ct, output_dir)

  plot_marker_dotplot(
    sub_obj,
    marker_table,
    annot_col = "cluster_subtipo",
    outfile   = file.path(output_dir, paste0("dotplot_markers_subclusters_", file_tag, ".pdf")),
    width = 18,
    height = 18
  )

  inspection_entries[[safe_name]] <- list(
    umap_plot = p_dim,
    obj       = sub_obj
  )
}

save_subcluster_composite(
  subcluster_list = inspection_entries,
  marker_table    = marker_table,
  output_dir      = output_dir,
  filename        = "subclustering_marker_inspection2.pdf"
)

# After reviewing the PDFs, manually decide whether to reassign subclusters.
# Nothing is changed in pbmc_harmony unless you run an apply_subcluster_reassignment()
# block intentionally.

# Example reassignment template, left commented on purpose:
#
# reassign <- list(
#   Epidermis.Hypocotyl.1 = c(
#     "0"      = "Epidermis Hypocotyl",
#     "1"      = "Epidermis Cotyledon",
#     "others" = "Epidermis Hypocotyl"
#   )
# )
#
# pbmc_harmony <- apply_subcluster_reassignment(
#   obj             = pbmc_harmony,
#   subcluster_list = subcluster_list,
#   reassign        = reassign,
#   source_col      = curation_col,
#   dest_col        = "celltype_curated"
# )
#
# save_pdf(
#   DimPlot(pbmc_harmony, group.by = "celltype_curated",
#           label = TRUE, repel = TRUE, raster = FALSE),
#   "umap_curated.pdf"
# )
# saveRDS(pbmc_harmony, file.path(dir_objects, "pbmc_harmony_curated.rds"))

# =============================================================================

message("\n✓ SECTION 11 COMPLETE: Subcluster marker inspection saved")
```


## Section 12 - EXPORT TO H5AD (Scanpy / Python)


If optional curation was skipped, this block copies `celltype_grouped` into `celltype_curated` before export.


```{r section_12_export_to_h5ad_scanpy_python}
# SECTION 12 — EXPORT TO H5AD (Scanpy / Python)
# =============================================================================
# Exports the curated object to AnnData h5ad format for Python-based
# trajectory and velocity analyses (Scanpy, scFates, Palantir — all
# pre-installed in the Docker image).

if (!"celltype_curated" %in% colnames(pbmc_harmony@meta.data)) {
  message("celltype_curated not found; using celltype_grouped as the final annotation for export.")
  pbmc_harmony$celltype_curated <- pbmc_harmony$celltype_grouped
  saveRDS(pbmc_harmony, file.path(dir_objects, "pbmc_harmony_curated.rds"))
}

export_to_scanpy(pbmc_harmony,
                 file.path(dir_objects, "pbmc_harmony_curated.h5ad"))

# To export a specific cell type:
# export_to_scanpy(
#   subset(pbmc_harmony, subset = celltype_curated == "Guard Cell"),
#   file.path(dir_objects, "GuardCell.h5ad")
# )


# =============================================================================

message("\n✓ SECTION 12 COMPLETE: Export to h5ad complete")
```
