---
title: "Results Report: WT vs pifq single-cell RNA-seq pipeline"
subtitle: "Executed results — real figures from a fresh run using the bibliography annotation (celltype); code shown as in the methods paper"
author: "SingleCell Pipeline"
date: "2026-07-28"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: pdflatex
    includes:
      in_header: latex_header.tex
    pandoc_args: ["--lua-filter=lang_box.lua"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,
  message = TRUE,
  warning = TRUE,
  error = FALSE,
  strip_comments = TRUE
)

strip_inline_comment <- function(line) {
  chars <- strsplit(line, "", fixed = TRUE)[[1]]
  in_single <- FALSE
  in_double <- FALSE
  escaped <- FALSE

  for (i in seq_along(chars)) {
    ch <- chars[[i]]
    if (escaped) {
      escaped <- FALSE
    } else if (ch == "\\") {
      escaped <- TRUE
    } else if (ch == "'" && !in_double) {
      in_single <- !in_single
    } else if (ch == '"' && !in_single) {
      in_double <- !in_double
    } else if (ch == "#" && !in_single && !in_double) {
      return(trimws(substr(line, 1, i - 1), which = "right"))
    }
  }
  line
}

source_hook <- knitr::knit_hooks$get("source")
knitr::knit_hooks$set(source = function(x, options) {
  if (isTRUE(options$strip_comments)) {
    lines <- strsplit(paste(x, collapse = "\n"), "\n", fixed = TRUE)[[1]]
    lines <- vapply(lines, strip_inline_comment, character(1), USE.NAMES = FALSE)
    lines <- lines[!grepl("^\\s*$", lines)]
    x <- paste(lines, collapse = "\n")
  }
  source_hook(x, options)
})
```

This methods-paper R Markdown documents the full WT vs pifq workflow: two
samples, `WT` and `pifq`, processed and integrated together, then carried
through three chapters — single-cell preprocessing, clustering and annotation
(Part 2); pseudobulk differential expression, GO and the co-expression network
(Part 3); and pseudotime trajectory analysis (Part 4). The code blocks are shown
exactly as in the methods paper (not executed inline); the figures below are the
**real outputs** of a fresh run of `workflow/step1_singlecell.R`,
`workflow/step2_degs.R` and `workflow/step3_pseudotime.py`, written to a separate
`resultados_rerun/` directory using the bibliography annotation (`celltype`).

The active annotation path is bibliography-based (`biblio_marks.txt` ->
`celltype_curated`). Reference-transfer annotation is not part of this draft.

## Minimum computing requirements {-}

The table below summarizes the computing requirements for running the full
pipeline on a personal computer, from raw FASTQ through the downstream Seurat/R
and Python workflow. These are not hard limits; below them the
pipeline still runs, just slower, and heavier steps (Cell Ranger, merging,
PCA/UMAP) risk stalling or exhausting memory.

| Component | Minimum | Recommended |
|---|---:|---:|
| CPU cores | 4 | 16 or more |
| RAM | 32 GB | 64-128 GB for merged multi-sample objects |
| Disk space | 80 GB free | 100 GB free |
| Container runtime | Any OCI-compatible runtime (e.g. Docker) | Docker or Singularity/Apptainer |

This workflow is packaged as a single public image (`matigara/scrnaseq:latest`,
on Docker Hub); the `docker-compose` YAML used to build and run it is kept in
the repository. Any container runtime able to run that image works; the setup
below covers both Docker and Singularity/Apptainer (the latter common on HPC
clusters, where Docker is usually not permitted).

Small pilot datasets can be run on a laptop or desktop if memory is sufficient.
For full experiments, repeated clustering sweeps, Cell Ranger processing from
FASTQ, or large merged Seurat objects, a server or external cluster is strongly
recommended. Cell Ranger is CPU- and memory-intensive and is not installed
inside the downstream `matigara/scrnaseq:latest` Docker image.

\newpage

# Part 1 - Environment and raw data

## Clone the repository

Choose or create a working directory, then clone the pipeline repository
into it:

```bash
mkdir -p ~/my-analysis
cd ~/my-analysis
git clone https://github.com/cliford2001/SingleCell.git
cd SingleCell
```

All paths referenced later in this document (`workflow/`) are relative to
this cloned repository. Its layout is:

```text
SingleCell/
|-- Dockerfile
|-- docker-compose.yml
|-- README.md
|-- data/
|   |-- AtTFDB_loci.txt
|   |-- biblio_marks.txt
|   |-- Arabidopsis_thaliana_TAIR10.1_Araport11_genome.fasta.gz
|   `-- Arabidopsis_thaliana_TAIR10.1_Araport11_annotation.gtf.gz
`-- workflow/
    |-- load_libraries.R                 # R packages
    |-- load_libraries_python.py         # Python packages
    |-- ScRNA_Analysis_Functions.R       # helper functions (Steps 1-2)
    |-- ScRNA_Pseudotime_Functions.py    # helper functions (Step 3)
    |-- step0_cellranger.sh              # Part 1 - Cell Ranger (bash)
    |-- step1_singlecell.R               # Part 2 - single-cell analysis
    |-- step2_degs.R                     # Part 3 - pseudobulk DE + GO
    `-- step3_pseudotime.py              # Part 4 - pseudotime (Python)
```

The bibliography marker table (`data/biblio_marks.txt`) is part of the
repository for this WT vs pifq run (see Section 1).

## Container startup (Docker or Singularity)

The container mounts the cloned repository itself as `/workspace`, so results
are written back into the repository (`resultados/`). Since results are not
pushed upstream, this is not a problem. The same public image runs under either
runtime; always pull it rather than rebuilding from the `Dockerfile`, so every
user gets the exact tested package versions.

**Docker.** Mount the cloned repository and open a shell:

```bash
docker run -it --rm \
  -v $(pwd):/workspace \
  matigara/scrnaseq:latest /bin/bash
```

Or, from inside the cloned repository, `docker compose run r` uses the bundled
`docker-compose.yml` (same image, same `/workspace` mount). On the current
server the Docker container is already running and the project is mounted at
`/workspace`.

**Singularity / Apptainer.** No root is required. Convert the public Docker Hub
image to a local `.sif` once, then open a shell binding the repository to
`/workspace`:

```bash
singularity build scrnaseq.sif docker://matigara/scrnaseq:latest
singularity exec --bind $(pwd):/workspace --pwd /workspace scrnaseq.sif /bin/bash
```

Apptainer imports the image environment, so the bundled Python (`/opt/venv`) and
R stay on `PATH`, and it runs as the invoking user, so files under `resultados/`
are owned by that user rather than by root. The `--pwd /workspace` flag starts
the shell inside the bound repository (otherwise Apptainer warns that the host
working directory does not exist in the container and falls back to `$HOME`).

## Raw data download and Cell Ranger processing

**Run this whole section outside the Docker container**, on a host, server, or
HPC where Cell Ranger is installed. Cell Ranger is *not* bundled inside the
`matigara/scrnaseq:latest` image (that image only covers the downstream R /
Python workflow), so running `cellranger` inside the container returns
`command not found`. Write the Cell Ranger outputs into the cloned repository
directory so they are visible from inside the container (which mounts that same
directory as `/workspace`).

**Install Cell Ranger 9.0.1** (proprietary 10x Genomics software; this workflow
uses version 9.0.1). Get it from the 10x Genomics *previous versions* page:
https://www.10xgenomics.com/support/software/cell-ranger/downloads/previous-versions

1. Log in and accept the license.
2. Find **Cell Ranger 9.0.1** and copy its `curl` command.
3. That URL is signed and time-limited (it carries an `Expires=` timestamp tied
   to your account), so paste the current one — it cannot be hard-coded here or
   reused later. A plain `git clone` of the source repo does *not* give a
   ready-to-run binary.

Then extract it and add it to your `PATH`:

```bash
# Paste the signed curl command for 9.0.1 from the 10x page, e.g.:
curl -o cellranger-9.0.1.tar.gz "https://cf.10xgenomics.com/releases/cell-exp/\
cellranger-9.0.1.tar.gz?Expires=...&Key-Pair-Id=...&Signature=..."

tar -xzvf cellranger-9.0.1.tar.gz
export PATH=$PWD/cellranger-9.0.1/bin:$PATH
cellranger --version
```

The single-cell workflow starts from Cell Ranger output, specifically the
`outs/filtered_feature_bc_matrix/` folder for each sample. If only raw FASTQ
files are available, first download or stage the FASTQs, then run Cell Ranger
for each sample before starting the R workflow.

The WT vs pifq FASTQ files used in this run are public data from
[Han et al., 2023, *Nature Plants*](https://www.nature.com/articles/s41477-023-01544-4#data-availability),
deposited at the CNCB Genome Sequence Archive under BioProject
[PRJCA016521](https://ngdc.cncb.ac.cn/bioproject/browse/PRJCA016521), run
accession `CRA010863`:

| Sample | Run accession | Files |
|---|---|---|
| ScWT | CRR775298 | `CRR775298_f1.fastq.gz`, `CRR775298_r2.fastq.gz` |
| Scpifq | CRR775297 | `CRR775297_f1.fastq.gz`, `CRR775297_r2.fastq.gz` |

Download the reads, then rename/symlink them to the naming Cell Ranger
expects (`<sample>_S1_L001_R1/R2_001.fastq.gz`):

```bash
# Run from the cloned repository root (paths are relative to it, not /workspace).
mkdir -p fastq/CRA010863

wget -c -P fastq/CRA010863 \
  ftp://download.big.ac.cn/gsa2/CRA010863/CRR775298/CRR775298_f1.fastq.gz
wget -c -P fastq/CRA010863 \
  ftp://download.big.ac.cn/gsa2/CRA010863/CRR775298/CRR775298_r2.fastq.gz
wget -c -P fastq/CRA010863 \
  ftp://download.big.ac.cn/gsa2/CRA010863/CRR775297/CRR775297_f1.fastq.gz
wget -c -P fastq/CRA010863 \
  ftp://download.big.ac.cn/gsa2/CRA010863/CRR775297/CRR775297_r2.fastq.gz

( cd fastq/CRA010863
  ln -s CRR775298_f1.fastq.gz ScWT_S1_L001_R1_001.fastq.gz
  ln -s CRR775298_r2.fastq.gz ScWT_S1_L001_R2_001.fastq.gz
  ln -s CRR775297_f1.fastq.gz Scpifq_S1_L001_R1_001.fastq.gz
  ln -s CRR775297_r2.fastq.gz Scpifq_S1_L001_R2_001.fastq.gz )
```

After FASTQs are available, run Cell Ranger for each sample. Cell Ranger itself
is not redistributed in this repository; download it from the official 10x
Genomics GitHub release
(https://github.com/10XGenomics/cellranger/releases/tag/cellranger-9.0.1) and
add it to your `PATH`. Run all commands in this section from the cloned
repository directory, on a host/server/HPC where Cell Ranger is installed (not
inside the container).

The reference genome (TAIR10.1) and Araport11 annotation ship compressed in the
repository under `data/`, so no download is needed. Decompress them and build
the Cell Ranger reference:

```bash
gunzip -k data/Arabidopsis_thaliana_TAIR10.1_Araport11_genome.fasta.gz
gunzip -k data/Arabidopsis_thaliana_TAIR10.1_Araport11_annotation.gtf.gz

cellranger mkref \
  --genome=Ara \
  --fasta=data/Arabidopsis_thaliana_TAIR10.1_Araport11_genome.fasta \
  --genes=data/Arabidopsis_thaliana_TAIR10.1_Araport11_annotation.gtf
```

Run Cell Ranger counts for each sample:

```bash
cellranger count --localcores=80 --id=ScWT --fastqs="$(pwd)/fastq/CRA010863" \
  --sample=ScWT --transcriptome=Ara --create-bam false

cellranger count --localcores=80 --id=Scpifq --fastqs="$(pwd)/fastq/CRA010863" \
  --sample=Scpifq --transcriptome=Ara --create-bam false
```

Use the resulting `filtered_feature_bc_matrix/` path in the sample manifest in
Section 1.

\newpage

# Part 2 - Single-cell analysis in R

From this point on, the pipeline runs in R inside the Docker container.
Start R:

```bash
R
```

The next two steps, Configuration and Initialization, set up the paths and
helper functions used by every section that follows.

## Configuration


Edit this block before running the pipeline.


```{r configuration}
# CONFIGURATION - Edit only this block before running the pipeline
# =============================================================================

# Directory containing the pipeline helper scripts.
# The cloned repository itself is mounted as /workspace.
PIPELINE_DIR <- "/workspace/workflow"

# Root directory for your project data and results.
# All result files will be written to DATA_DIR/resultados/<step>/
DATA_DIR   <- "/workspace/."
base_dir   <- file.path(DATA_DIR, "resultados")

# =============================================================================
```


## Initialization


Load helper scripts, set reproducibility options, and create output directories.


```{r initialization}
# INITIALIZATION
# =============================================================================

# -- Load helper scripts --------------------------------------------------------
# Each file is fully documented at https://github.com/cliford2001/SingleCell
source(file.path(PIPELINE_DIR, "load_libraries.R"))          # all R packages
source(file.path(PIPELINE_DIR, "ScRNA_Analysis_Functions.R"))# analysis functions

set.seed(1807)
options(Seurat.allow.s4 = FALSE)  # required for Seurat 5 compatibility
setwd(DATA_DIR)

# -- Create per-step output directories ----------------------------------------
list2env(create_pipeline_dirs(base_dir), envir = .GlobalEnv)  # creates output folders and loads their paths as variables

# output_dir is the global variable used by save_pdf / save_qc / save_vln helpers.
# It is reassigned at the start of each section to the appropriate step directory.
output_dir <- base_dir


# =============================================================================
# ########################  PART 1 - SINGLE-CELL ANALYSIS  ####################
# =============================================================================
```

\newpage

## Section 1 - Data loading and pre-filter QC

```{r section_01_data_loading_and_pre_filter_qc}
# SECTION 1 - DATA LOADING AND PRE-FILTER QC
# =============================================================================
# Each sample is loaded from its input file and mitochondrial / chloroplast
# read fractions are computed per cell to guide filtering thresholds in
# Section 2. Pre-filter violin plots are saved to 01_qc/.
#
# +- CHANGE FOR YOUR ORGANISM --------------------------------------------------
#   Arabidopsis : mt_pattern = "^ATMG"  |  cp_pattern = "^ATCG"
#   Human       : mt_pattern = "^MT-"   |  cp_pattern = NULL
#   Mouse       : mt_pattern = "^mt-"   |  cp_pattern = NULL
# +-----------------------------------------------------------------------------
output_dir <- dir_01

# -- Sample manifest (CellRanger filtered_feature_bc_matrix) ------------------
# Add one entry per sample. Each entry needs:
#   file      - path to the filtered_feature_bc_matrix/ directory (relative to DATA_DIR)
#   label     - unique name for this sample (appears in all plots)
#   condition - experimental group this sample belongs to
samples <- list(
  list(
    file      = "ScWT/outs/filtered_feature_bc_matrix",
    label     = "WT",
    condition = "WT"
  ),
  list(
    file      = "Scpifq/outs/filtered_feature_bc_matrix",
    label     = "pifq",
    condition = "pifq"
  )
)


# -- Plot colors (one color per sample label) -----------------------------------
colors <- c(
  "WT"  = "#66c2a5",
  "pifq"  = "#fc8d62"
)


mt_pattern <- "^ATMG"  # Arabidopsis mitochondrial genes
cp_pattern <- "^ATCG"  # Arabidopsis chloroplast genes


# Load all samples using the helper function
seurat_list_raw <- load_seurat_samples(samples = samples,
                                       DATA_DIR = DATA_DIR,
                                       mt_pattern = mt_pattern,
                                       cp_pattern = cp_pattern)

plot_qc_batch(seurat_list_raw, colors, "qc_prefilter.pdf")


# =============================================================================

message("\nOK SECTION 1 COMPLETE: QC pre-filter plots saved")
```

Representative output from Section 1:

![Pre-filter QC violin plots.](rerun_figs/qc_prefilter.pdf){width=30%}

\newpage

## Section 2 - Cell filtering and doublet detection

```{r section_02_cell_filtering_and_doublet_detection}
output_dir <- dir_01

seurat_list <- filter_seurat_samples(seurat_list_raw, min_features = 200, max_mt = 5)

plot_qc_batch(seurat_list, colors, "qc_postfilter.pdf")

saveRDS(seurat_list, file.path(dir_objects, "seurat_list_postfilter.rds"))

message("\nOK SECTION 2 COMPLETE: filtering and doublet detection complete")
```

Representative output from Section 2:

![Post-filter QC violin plots.](rerun_figs/qc_postfilter.pdf){width=30%}

\newpage

## Section 3 - Merge and initial preprocessing

```{r section_03_merge_and_initial_preprocessing}
output_dir <- dir_01

pbmc_harmony <- reduce(seurat_list, merge) %>%
  NormalizeData(verbose = FALSE) %>%
  FindVariableFeatures(
    selection.method = "vst", nfeatures = 2000, verbose = FALSE
  ) %>%
  ScaleData(verbose = FALSE) %>%
  RunPCA(npcs = 30, verbose = FALSE) %>%
  RunUMAP(reduction = "pca", dims = 1:30, verbose = FALSE)

save_pdf(
  DimPlot(pbmc_harmony, group.by = "orig.ident", cols = colors),
  "umap_preharmony.pdf"
)

saveRDS(pbmc_harmony, file.path(dir_objects, "pbmc_harmony_preharmony.rds"))

message("\nOK SECTION 3 COMPLETE: merge and preprocessing complete")
```

Representative output from Section 3:

![UMAP before Harmony correction, colored by sample.](rerun_figs/umap_preharmony.pdf){width=30%}

\newpage

## Section 4 - Harmony batch correction

```{r section_04_harmony_batch_correction}
output_dir <- dir_01

pbmc_harmony <- pbmc_harmony %>%
  RunHarmony("orig.ident", plot_convergence = FALSE) %>%
  RunUMAP(reduction = "harmony", dims = 1:30, verbose = FALSE)

save_pdf(
  DimPlot(pbmc_harmony, group.by = "orig.ident", cols = colors),
  "umap_postharmony.pdf"
)

saveRDS(pbmc_harmony, file.path(dir_objects, "pbmc_harmony_postharmony.rds"))

message("\nOK SECTION 4 COMPLETE: Harmony integration complete")
```

Representative output from Section 4:

![UMAP after Harmony correction, colored by sample.](rerun_figs/umap_postharmony.pdf){width=30%}

\newpage

## Section 5 - Resolution optimization

```{r section_05_resolution_optimization}
resolutions_test <- c(0.15, 0.35, 0.45, 0.55, 1.0)

output_dir <- dir_02

k_range <- 1:31
pca_data <- Embeddings(pbmc_harmony, "pca")[, 1:30]
wss <- sapply(
  k_range,
  function(k) kmeans(pca_data, centers = k, nstart = 4)$tot.withinss
)

elbow_plot <- ggplot(data.frame(k = k_range, wss = wss), aes(k, wss)) +
  geom_line() +
  geom_point() +
  labs(
    x = "Number of clusters (k)",
    y = "Within-cluster sum of squares"
  ) +
  theme_minimal()

save_pdf(elbow_plot, "elbow_plot.pdf", w = 18, h = 18)

clu <- pbmc_harmony %>%
  RunUMAP(reduction = "harmony", dims = 1:30, verbose = FALSE) %>%
  FindNeighbors(reduction = "harmony", dims = 1:30, k.param = 20, verbose = FALSE)

for (res in resolutions_test)
  clu <- FindClusters(clu, resolution = res, algorithm = 4, verbose = FALSE)

save_pdf(clustree(clu, prefix = "RNA_snn_res."), "clustree2.pdf", w = 18, h = 18)

message("\nOK SECTION 5 COMPLETE: elbow plot and clustree saved")
```

Representative outputs from Section 5:

```{=latex}
\begin{center}
```
![Elbow diagnostic used to inspect candidate cluster numbers.](rerun_figs/elbow_plot.pdf){width=30%} ![Cluster tree across candidate Leiden resolutions.](rerun_figs/clustree.pdf){width=30%}
```{=latex}
\end{center}
```

\newpage

## Section 6 - Final clustering

```{r section_06_final_clustering}
cluster_resolution <- 0.35
output_dir <- dir_02

pbmc_harmony <- pbmc_harmony %>%
  RunUMAP(reduction = "harmony", dims = 1:30, verbose = FALSE) %>%
  FindNeighbors(reduction = "harmony", dims = 1:30, k.param = 20, verbose = FALSE) %>%
  FindClusters(resolution = cluster_resolution, algorithm = 4, verbose = FALSE)

Idents(pbmc_harmony) <- "seurat_clusters"
save_pdf(
  DimPlot(pbmc_harmony, group.by = "seurat_clusters", label = TRUE),
  "umap_seuratclusters.pdf"
)

message("\nOK SECTION 6 COMPLETE: final clustering complete")
```

Representative output from Section 6:

![Final UMAP colored by numbered Seurat clusters.](rerun_figs/umap_seuratclusters.pdf){width=30%}

\newpage

## Section 7 - Cell-type annotation

```{r section_07_cell_type_annotation}
output_dir <- dir_03

biblio_marks_file <- file.path(DATA_DIR, "data", "biblio_marks.txt")
marker_table <- read.table(
  biblio_marks_file, header = TRUE, sep = "\t", quote = ""
)

markers <- find_markers(
  pbmc_harmony,
  output_file = file.path(output_dir, "FindAllMarkers.tsv")
)

pbmc_harmony <- annotate_by_markers(
  pbmc_harmony, markers,
  reference_file = biblio_marks_file
)

plot_marker_dotplot(
  pbmc_harmony,
  marker_table,
  annot_col = "celltype",
  outfile   = file.path(
    output_dir, "dotplot_marker_table_annotation_biblio.pdf"
  ),
  width = 18, height = 18
)

save_pdf(
  DimPlot(
    pbmc_harmony, group.by = "celltype",
    label = TRUE, repel = TRUE, raster = FALSE
  ),
  "umap_annotation_biblio.pdf"
)

message("\nOK SECTION 7 COMPLETE: cell-type annotation complete")
```

Representative outputs from Section 7:

```{=latex}
\begin{center}
```
![Marker dotplot after bibliography-based cell-type annotation.](rerun_figs/dotplot_annotation.pdf){width=30%} ![UMAP colored by bibliography-based cell-type annotation.](rerun_figs/umap_annotation.pdf){width=30%}
```{=latex}
\end{center}
```

\newpage

## Section 8 - Annotated clustree

```{r section_08_annotated_clustree}
output_dir <- dir_03

Mode <- function(x) {
  x <- as.character(x)
  x <- x[!is.na(x) & nzchar(x)]
  if (length(x) == 0) return(NA_character_)
  names(sort(table(x), decreasing = TRUE))[1]
}

stopifnot(exists("clu"))
stopifnot("celltype" %in% colnames(pbmc_harmony@meta.data))

celltype_label <- as.character(pbmc_harmony$celltype)
names(celltype_label) <- Cells(pbmc_harmony)
clu$celltype_label <- celltype_label[Cells(clu)]

print(table(clu$celltype_label, useNA = "ifany"))

save_pdf(
  clustree(
    clu, prefix = "RNA_snn_res.",
    node_label = "celltype_label", node_label_aggr = "Mode"
  ),
  "clustree_annotated.pdf", w = 14, h = 14
)

message("\nOK SECTION 8 COMPLETE: annotated clustree saved")
```

Representative output from Section 8:

![Cluster tree labeled by the dominant bibliography-based annotation in each node.](rerun_figs/clustree_annotated.pdf){width=30%}

\newpage

## Section 9 - Gene expression visualization

```{r section_09_gene_expression_visualization}
gene <- "AT5G26000"

output_dir <- dir_04

pbmc_harmony <- JoinLayers(pbmc_harmony)
Idents(pbmc_harmony) <- "celltype"

save_vln(
  VlnPlot(pbmc_harmony, features = gene),
  "vln_gene_all.pdf"
)

save_pdf(
  FeaturePlot(pbmc_harmony, features = gene),
  "feature_gene_all.pdf"
)

message("\nOK SECTION 9 COMPLETE: gene expression visualization saved")
```

Representative outputs from Section 9:

```{=latex}
\begin{center}
```
![Violin plot for the selected gene across all annotated cell types.](rerun_figs/vln_gene.pdf){width=30%} ![UMAP expression overlay for the selected gene.](rerun_figs/feature_gene.pdf){width=30%}
```{=latex}
\end{center}
```

\newpage

## Section 10 - Cell-type grouping [optional]

```{r section_10_cell_type_grouping_optional, strip_comments=FALSE}
grouping <- c()
# Example:
# grouping <- c(
#   "Epidermis Hypocotyl.1" = "Epidermis Hypocotyl",
#   "Epidermis Hypocotyl.2" = "Epidermis Hypocotyl"
# )

output_dir <- dir_05

if (length(grouping) > 0) {
  pbmc_harmony$celltype_grouped <- recode(pbmc_harmony$celltype, !!!grouping)
} else {
  pbmc_harmony$celltype_grouped <- pbmc_harmony$celltype
}

save_pdf(
  DimPlot(
    pbmc_harmony, group.by = "celltype_grouped",
    label = TRUE, repel = TRUE, raster = FALSE
  ),
  "umap_grouped.pdf"
)

message("\nOK SECTION 10 COMPLETE: cell-type grouping complete")
```

Representative output from Section 10:

![UMAP colored by grouped cell-type labels.](rerun_figs/umap_grouped.pdf){width=30%}

\newpage

## Section 11 - Subcluster marker inspection [optional]

This section is intentionally interactive. `inspect_subcluster_markers()`
subclusters one cluster and saves a combined figure (subcluster UMAP, marker
dotplot, and a small FeaturePlot grid) to help decide whether curation is
needed. After reviewing the figures, `curate_clusters()` re-inspects the chosen
clusters and reassigns their subclusters to the labels you provide, writing the
result into `celltype_curated`. Both helpers live in
`ScRNA_Analysis_Functions.R`.

```{r section_11_interactive_cell_type_curation_optional, eval=FALSE, strip_comments=FALSE}
output_dir <- dir_05
Idents(pbmc_harmony) <- "celltype"

# 1. Inspect a cluster (saves the combined subcluster + marker figure)
inspect_subcluster_markers(
  pbmc_harmony, cluster_id = "2",
  marker_table = marker_table, output_dir = output_dir
)

# 2. Map each cluster's subclusters to labels ("others" = any ID not listed),
#    then curate them in one call. Example for two clusters:
reassign <- list(
  "2" = c("0" = "Identity A", "others" = "Unresolved"),
  "4" = c("1" = "Identity B", "others" = "Unresolved")
)
pbmc_harmony <- curate_clusters(
  pbmc_harmony, reassign,
  marker_table = marker_table, output_dir = output_dir
)

# 3. Save the curated annotation UMAP
save_pdf(
  DimPlot(pbmc_harmony, group.by = "celltype_curated",
          label = TRUE, repel = TRUE, raster = FALSE),
  "umap_curated.pdf"
)

message("\nOK SECTION 11 COMPLETE: curation complete")
```

*Section 11 (interactive subcluster curation) was not re-executed in this run.
All downstream analyses use the bibliography annotation (`celltype`) directly,
so unresolved clusters keep their numeric labels (`2`, `4`, `5`, `7`).*

## Section 12 - Export the curated object

Closes Chapter 1 by saving the curated object in two formats: an `.rds`
checkpoint for the R-based Chapter 2, and an AnnData `.h5ad` for the
Python-based Chapter 3 trajectory and velocity analyses (Scanpy, scFates,
Palantir - all pre-installed in the Docker image).

```{r section_12_export, strip_comments=FALSE}
# Checkpoint - restore with:
#   pbmc_harmony <- readRDS(file.path(dir_objects, "pbmc_harmony_curated.rds"))
saveRDS(pbmc_harmony, file.path(dir_objects, "pbmc_harmony_curated.rds"))

# Export the curated object to AnnData h5ad format for Python-based
# trajectory and velocity analyses.
export_to_scanpy(
  pbmc_harmony,
  file.path(dir_objects, "pbmc_harmony_curated.h5ad")
)

# To export a specific cell type only:
# export_to_scanpy(
#   subset(pbmc_harmony, subset = celltype_curated == "guard cell"),
#   file.path(dir_objects, "GuardCell.h5ad")
# )

message("\nOK SECTION 12 COMPLETE: curated object saved (.rds + .h5ad)")
```

\newpage

# Part 3 - Pseudobulk differential expression and GO

This part continues from the object saved at the end of Chapter 1. It
aggregates cells into pseudo-replicates per cell type, runs pseudobulk DESeq2
for the WT vs pifq contrast, and summarizes results with volcano plots,
differential tables, GO enrichment, a log2FC heatmap, and co-expression
networks.

```{r part3_load_curated}
pbmc_harmony <- readRDS(file.path(dir_objects, "pbmc_harmony_curated.rds"))
```

## Section 13 - Cell-type subsets

The annotation column used to define cell types is a user choice. Depending on
how Chapter 1 was run, this is one of `celltype` (bibliography annotation),
`celltype_grouped` (broad groups from Section 10), or `celltype_curated`
(manual curation from Section 11). All downstream sections in this part use
whichever column is set here.

```{r section_13_cell_type_subsets}
pseudobulk_annot_col <- "celltype_curated"

# Inspect how many cells each cell type has before subsetting
table(pbmc_harmony[[pseudobulk_annot_col]])

cell_type_subsets <- create_cell_type_subsets(
  pbmc_harmony, annot_col = pseudobulk_annot_col
)

message("\nOK SECTION 13 COMPLETE: cell-type subsets created")
```

Cells per cell type (first three shown):

| Cell type | Cells |
|---|---:|
| Epidermis_Hypocotyl_1 | 1847 |
| Epidermis_Cotyledon | 1461 |
| Mesophyll | 488 |

## Section 14 - Pseudo-replicate assignment

```{r section_14_pseudoreplicate_assignment}
pseudobulk_conditions <- NULL
n_pseudoreps          <- 3

cell_type_subsets_replicates <- assign_pseudoreplicates_batch(
  cell_type_subsets,
  pseudobulk_conditions = pseudobulk_conditions,
  n_pseudoreps          = n_pseudoreps
)

# Inspect cells per pseudo-replicate for one cell type
table(cell_type_subsets_replicates[["Mesophyll"]]$replicate)

message("\nOK SECTION 14 COMPLETE: pseudo-replicates assigned")
```

Cell types present in only one condition (e.g. `C2`, present in only 57 cells
from a single sample) are dropped here, since a WT vs pifq contrast needs both
conditions represented.

Cells per pseudo-replicate (first three cell types shown):

| Cell type | Replicate | Cells |
|---|---|---:|
| Epidermis_Hypocotyl_1 | pifq_rep1 | 329 |
| Epidermis_Hypocotyl_1 | pifq_rep2 | 366 |
| Epidermis_Hypocotyl_1 | pifq_rep3 | 346 |
| Epidermis_Hypocotyl_1 | WT_rep1 | 279 |
| Epidermis_Hypocotyl_1 | WT_rep2 | 272 |
| Epidermis_Hypocotyl_1 | WT_rep3 | 255 |
| Epidermis_Cotyledon | pifq_rep1 | 316 |
| ... | ... | ... |

## Section 15 - Pseudobulk tables and DESeq2

```{r section_15_pseudobulk_deseq2}
comparisons <- list(
  list(conds = c("WT", "pifq"), tag = "WT_vs_pifq")
)
cell_types_to_analyze <- NULL
output_dir <- dir_06

deseq2_results <- run_pseudobulk_deseq2_analysis(
  cell_type_subsets_replicates = cell_type_subsets_replicates,
  comparisons    = comparisons,
  output_dir     = output_dir,
  cell_types     = cell_types_to_analyze,
  pseudobulk_dir = file.path(dir_objects, "pseudobulk_replicas")
)

message("\nOK SECTION 15 COMPLETE: pseudobulk aggregation and DESeq2 complete")
```

One `DESeq2_<cell_type>_WT_vs_pifq.csv` is written per cell type: the standard
DESeq2 results table (one row per gene, with `log2FoldChange`, `pvalue`, and
`padj`). Positive log2FC means higher expression in pifq. First rows of the
Mesophyll table:

```text
             baseMean  log2FoldChange   lfcSE      stat     pvalue      padj
AT1G01010    46.22     -2.102           0.307     -6.857   7.0e-12     9.7e-11
AT1G01020    48.78      0.100           0.326      0.306   7.6e-01     8.5e-01
AT1G01030   172.27      0.437           0.306      1.428   1.5e-01     2.8e-01
```

## Section 16 - Volcano plots

```{r section_16_volcano_plots}
volcano_tag <- "WT_vs_pifq"
padj_cut    <- 0.05
lfc_cut     <- 1

render_volcano_plots(
  results_dir = file.path(dir_06, volcano_tag),
  output_dir  = file.path(dir_06, volcano_tag, "volcano"),
  pdf_name    = paste0("VolcanoPlots_", volcano_tag, ".pdf"),
  padj_cut    = padj_cut,
  lfc_cut     = lfc_cut
)

message("\nOK SECTION 16 COMPLETE: volcano plots saved")
```

Representative output from Section 16:

![Volcano plot for Mesophyll, WT vs pifq.](rerun_figs/volcano.pdf){width=30%}

## Section 17 - Differential gene tables

```{r section_17_differential_gene_tables}
diff_prefix <- paste0("diff_table_", volcano_tag)

diff_tables <- build_differential_tables(
  results_dir = file.path(dir_06, volcano_tag),
  output_dir  = file.path(dir_06, volcano_tag),
  padj_cut    = padj_cut,
  lfc_cut     = lfc_cut,
  prefix      = diff_prefix
)

message("\nOK SECTION 17 COMPLETE: differential gene tables saved")
```

This writes two combined matrices (genes x cell types). The class matrix
(`diff_table_WT_vs_pifq_*.tsv`) uses `+1`/`-1`/`0` to flag genes up-, down-, or
not significantly regulated in each cell type; the log2FC matrix
(`tabla_log2FC_*.tsv`) holds the actual fold changes (`NA` where not
significant). Example rows for three annotated cell types (Epidermis
Cotyledon, guard cell, Mesophyll):

```text
# class matrix                       # log2FC matrix
gene_id    EpiCot  guard  Meso       gene_id    EpiCot  guard  Meso
AT1G29930    1       1      1        AT1G29930    4.79    3.51   2.02
AT1G54410   -1       1     -1        AT1G54410   -1.23    2.52  -1.88
AT1G55670    1       1      1        AT1G55670    3.58    4.05   1.66
```

## Section 18 - GO enrichment

```{r section_18_go_enrichment}
go_space    <- "BP"
go_orgdb    <- org.At.tair.db
go_keytype  <- "TAIR"
padj_cutoff <- 0.05

go_results <- run_go_enrichment_for_contrast(
  results_dir  = file.path(dir_06, volcano_tag),
  output_dir   = file.path(dir_07, volcano_tag),
  orgdb        = go_orgdb,
  keytype      = go_keytype,
  go_space     = go_space,
  padj_cutoff  = padj_cutoff,
  contrast_tag = volcano_tag
)

message("\nOK SECTION 18 COMPLETE: GO enrichment complete")
```

Representative output from Section 18:

![GO enrichment (BP) for Mesophyll, WT vs pifq.](rerun_figs/go_mesophyll.pdf){width=30%}

## Section 19 - Log2FC heatmap

Rows (genes) are ordered by hierarchical clustering, so genes with similar
cross-cell-type profiles group together. Columns (cell types) follow the
marker-table order via `cell_order`; cell types not present in the marker
table (the unannotated clusters) are placed at the end.

```{r section_19_log2fc_heatmap}
heatmap_limits <- c(-5, 5)

build_logfc_heatmap(
  logfc_table  = diff_tables$logfc,
  contrast_tag = volcano_tag,
  output_dir   = file.path(dir_06, volcano_tag),
  limits       = heatmap_limits,
  cell_order   = unique(marker_table$cell.types)
)

message("\nOK SECTION 19 COMPLETE: log2FC heatmap saved")
```

Representative output from Section 19:

![log2FC heatmap: rows hierarchically clustered, columns in marker-table order (unannotated clusters last), WT vs pifq.](rerun_figs/heatmap.pdf){width=30%}

\newpage

## Section 20 - Gene-gene coexpression and TF network

`run_tf_coexpression_network()` does the whole network step in one call: it
builds the hdWGCNA gene-gene co-expression network from the significant DE
genes, exports the edge list (`edges_unified.tsv`, `source | target | weight`),
then filters to TF-containing pairs (AtTFDB) and draws the TF network colored
by DE direction. No module-clustering plots are produced. Only the parameters
below are meant to be edited.

```{r section_20_tf_coexpression_network}
run_tf_coexpression_network(
  seurat_obj      = pbmc_harmony,
  de_table_path   = file.path(dir_06, volcano_tag, "tabla_log2FC_fc1_padj_005.tsv"),
  tf_list_path    = file.path(DATA_DIR, "data", "AtTFDB_loci.txt"),
  output_dir      = dir_08,
  annot_col       = pseudobulk_annot_col,
  contrast_tag    = volcano_tag,
  n_metacells     = 50,     # metacells per cell type x sample
  min_module_size = 20,     # minimum genes per co-expression module
  deep_split      = 2,      # module-splitting sensitivity (0-4)
  soft_power      = NULL,   # NULL = auto-detect the soft power
  tom_threshold   = 0.2,    # minimum co-expression weight to keep an edge
  n_hub_label     = 15      # number of top hub TFs to label
)

message("\nOK SECTION 20 COMPLETE: TF coexpression network saved")
```

Representative output from Section 20:

![TF co-expression network, colored by DE direction.](rerun_figs/network.pdf){width=30%}

\newpage

# Part 4 - Pseudotime trajectory analysis (Python)

Chapter 3 runs in Python (scanpy, Palantir, scFates), starting from the h5ad
object exported at the end of Chapter 1
(`resultados/objects/pbmc_harmony_curated.h5ad`). Exit R and start Python
inside the same Docker container:

```bash
q()          # leave R (do not save the workspace)
python3
```

## Section 23 - Setup

```{python section_23_setup, eval=FALSE}
import os

PIPELINE_DIR = "/workspace/workflow"
DATA_DIR     = "/workspace/."
base_dir     = os.path.join(DATA_DIR, "resultados")

exec(open(os.path.join(PIPELINE_DIR, "load_libraries_python.py")).read(), globals())
exec(open(os.path.join(PIPELINE_DIR, "ScRNA_Pseudotime_Functions.py")).read(), globals())

dir_pseudotime = os.path.join(base_dir, "09_pseudotime")
os.makedirs(dir_pseudotime, exist_ok=True)

print("SECTION 23 COMPLETE: setup done")
```

## Section 24 - Load curated object

Reads the AnnData object, converts the Seurat-style UMAP/PCA/HARMONY coordinate
names to scanpy names, saves an overview UMAP, and prints the available cell
types (used to choose the trajectory in Section 25).

```{python section_24_load_object, eval=FALSE}
INPUT_H5AD     = "/workspace/resultados/objects/pbmc_harmony_curated.h5ad"
ANNOTATION_COL = "celltype_curated"
N_JOBS         = 4

adata, N_JOBS = load_curated_object(
    input_h5ad     = INPUT_H5AD,
    dir_pseudotime = dir_pseudotime,
    annotation_col = ANNOTATION_COL,
    n_jobs         = N_JOBS,
)
```

Representative output from Section 24:

![Overview UMAP colored by curated cell type.](rerun_figs/pt_overview.pdf){width=30%}

## Section 25 - Cell type selection

Picks the cell types used to build the trajectory. Here we follow the stomatal
lineage: cotyledon epidermis differentiating into guard cells. Names must match
the Section 24 list exactly.

```{python section_25_cell_selection, eval=FALSE}
TRAJECTORY_CLUSTERS = ["Epidermis Cotyledon", "guard cell"]

adata_sub = preview_trajectory_selection(
    adata          = adata,
    clusters       = TRAJECTORY_CLUSTERS,
    annotation_col = ANNOTATION_COL,
    dir_pseudotime = dir_pseudotime,
)
```

Representative output from Section 25:

![UMAP of the cells selected for the trajectory.](rerun_figs/pt_selection.pdf){width=30%}

## Section 26 - Trajectory inference

Builds the trajectory with scFates: Palantir diffusion maps, a principal tree,
root selection at the progenitor cluster, and pseudotime assignment. Add more
`trajectory_run(...)` entries to sweep parameters.

```{python section_26_trajectory, eval=FALSE}
ROOT_CLUSTER = "Epidermis Cotyledon"

TRAJECTORY_RUNS = [
    trajectory_run(nodes=50, sigma=0.1, lambda_value=200, eigs=8, seed=3),
]

adata_traj, selected_trajectory_dir, trajectory_runs = run_trajectory_runs(
    adata           = adata,
    clusters        = TRAJECTORY_CLUSTERS,
    root_cluster    = ROOT_CLUSTER,
    annotation_col  = ANNOTATION_COL,
    output_base_dir = dir_pseudotime,
    runs            = TRAJECTORY_RUNS,
)
```

Representative outputs from Section 26:

```{=latex}
\begin{center}
```
![Trajectory graph colored by cell type.](rerun_figs/pt_traj_annotation.pdf){width=30%} ![Pseudotime overlaid on the force-directed layout.](rerun_figs/pt_traj_time.pdf){width=30%}
```{=latex}
\end{center}
```

## Section 27 - Plot genes on trajectory

Plots selected genes on the trajectory layout.

```{python section_27_gene_plots, eval=FALSE}
GENES = ["AT3G24140", "AT4G21750", "AT4G20260"]
name  = os.path.basename(selected_trajectory_dir)

ensure_expression_in_X(adata_traj)

gene_plots_dir = os.path.join(selected_trajectory_dir, "gene_plots")
os.makedirs(gene_plots_dir, exist_ok=True)

fig = sc.pl.draw_graph(
    adata_traj, color=GENES, title=GENES, show=False, return_fig=True
)
save_plot_18x18(fig, os.path.join(gene_plots_dir, f"{name}_gene_plots.pdf"))
plt.close(fig)

print("SECTION 27 COMPLETE: gene trajectory plots saved")
```

Representative output from Section 27:

![Selected genes on the trajectory layout.](rerun_figs/pt_gene_plots.pdf){width=30%}

## Section 28 - Gene trends

Fits gene expression along pseudotime and saves two heatmaps: the top variable
genes, and a custom highlighted set.

```{python section_28_gene_trends, eval=FALSE}
TOP_N           = 10
HIGHLIGHT_GENES = ["AT3G24140", "AT4G21750", "AT4G20260"]
ORDERING        = "max"
SPLINE_DF       = 3

top_path, highlight_path = run_step29_gene_trends(
    adata        = adata_traj,
    run_dir      = selected_trajectory_dir,
    name         = os.path.basename(selected_trajectory_dir),
    custom_genes = HIGHLIGHT_GENES,
    top_n        = TOP_N,
    ordering     = ORDERING,
    n_jobs       = N_JOBS,
    spline_df    = SPLINE_DF,
)

print("SECTION 28 COMPLETE: gene trend plots saved")
```

Representative outputs from Section 28 — the top variable genes and the
highlighted gene list:

```{=latex}
\begin{center}
```
![Top 10 variable genes along pseudotime.](rerun_figs/pt_gene_trends.pdf){width=48%} ![Highlighted gene list along pseudotime.](rerun_figs/pt_gene_trends_highlight.pdf){width=48%}
```{=latex}
\end{center}
```

## Section 29 - Final export

Saves the fully processed pseudotime object (trajectory tree, pseudotime, and
force-directed layout) as the pipeline's final Chapter 3 output, alongside the
curated Chapter 1 object in `resultados/objects/`.

```{python section_29_final_export, eval=FALSE}
final_h5ad = os.path.join(base_dir, "objects", "pbmc_pseudotime_final.h5ad")
os.makedirs(os.path.dirname(final_h5ad), exist_ok=True)
adata_traj.write_h5ad(final_h5ad)

print(f"SECTION 29 COMPLETE: final pseudotime object saved -> {final_h5ad}")
```
```{=latex}
\newpage
```

# Supplementary Table — Software dependencies and versions {-}

Package versions are those bundled in the `matigara/scrnaseq:latest` Docker image
(the public image used for all R and Python steps). R **base-priority** packages
are omitted. Cell Ranger is 10x Genomics proprietary software run *outside* the
container (Part 1); its version is fixed by the workflow.

## Interpreters and command-line tools {-}

| Tool | Version |
|:-----|:--------|
| Cell Ranger | 9.0.1 |
| R | 4.5.3 |
| Python | 3.12.3 |


## R packages (421) {-}


```{=latex}
\footnotesize
```
| Package | Version | Package | Version |
|:--------|:--------|:--------|:--------|
| `abind` | 1.4-8 | `locfit` | 1.5-9.12 |
| `anndataR` | 1.0.2 | `lubridate` | 1.9.5 |
| `AnnotationDbi` | 1.72.0 | `magrittr` | 2.0.5 |
| `ape` | 5.8-1 | `maps` | 3.4.3 |
| `aplot` | 0.2.9 | `MASS` | 7.3-65 |
| `askpass` | 1.2.1 | `Matrix` | 1.7-5 |
| `assertthat` | 0.2.1 | `MatrixGenerics` | 1.22.0 |
| `assorthead` | 1.4.0 | `MatrixModels` | 0.5-4 |
| `backports` | 1.5.1 | `matrixStats` | 1.5.0 |
| `base64enc` | 0.1-6 | `memoise` | 2.0.1 |
| `basilisk` | 1.22.0 | `mgcv` | 1.9-4 |
| `batchelor` | 1.26.0 | `microbenchmark` | 1.5.0 |
| `beachmat` | 2.26.0 | `mime` | 0.13 |
| `beeswarm` | 0.4.0 | `miniUI` | 0.1.2 |
| `BH` | 1.90.0-1 | `minqa` | 1.2.8 |
| `biglm` | 0.9-3 | `modelr` | 0.1.11 |
| `Biobase` | 2.70.0 | `monocle3` | 1.4.27 |
| `BiocGenerics` | 0.56.0 | `nlme` | 3.1-169 |
| `BiocManager` | 1.30.27 | `nloptr` | 2.2.1 |
| `BiocNeighbors` | 2.4.0 | `nnet` | 7.3-20 |
| `BiocParallel` | 1.44.0 | `numDeriv` | 2016.8-1.1 |
| `BiocSingular` | 1.26.1 | `openssl` | 2.4.1 |
| `BiocVersion` | 3.22.0 | `org.At.tair.db` | 3.22.0 |
| `Biostrings` | 2.78.0 | `otel` | 0.2.0 |
| `bit` | 4.6.0 | `pak` | 0.9.5 |
| `bit64` | 4.8.2 | `parallelly` | 1.47.0 |
| `bitops` | 1.0-9 | `patchwork` | 1.3.2 |
| `blob` | 1.3.0 | `pbapply` | 1.7-4 |
| `boot` | 1.3-32 | `pbkrtest` | 0.5.5 |
| `BPCells` | 0.3.1 | `pbmcapply` | 1.5.1 |
| `brew` | 1.0-10 | `pheatmap` | 1.0.13 |
| `brio` | 1.1.5 | `pillar` | 1.11.1 |
| `broom` | 1.0.13 | `pkgbuild` | 1.4.8 |
| `bslib` | 0.11.0 | `pkgconfig` | 2.0.3 |
| `cachem` | 1.1.0 | `pkgdown` | 2.2.0 |
| `Cairo` | 1.7-0 | `pkgload` | 1.5.2 |
| `callr` | 3.7.6 | `plotly` | 4.12.0 |
| `car` | 3.1-5 | `plyr` | 1.8.9 |
| `carData` | 3.0-6 | `png` | 0.1-9 |
| `caTools` | 1.18.3 | `polyclip` | 1.10-7 |
| `cellranger` | 1.1.0 | `polynom` | 1.4-1 |
| `checkmate` | 2.3.4 | `praise` | 1.0.0 |
| `circlize` | 0.4.18 | `preprocessCore` | 1.72.0 |
| `class` | 7.3-23 | `prettyunits` | 1.2.0 |
| `classInt` | 0.4-11 | `processx` | 3.9.0 |
| `cli` | 3.6.6 | `profvis` | 0.4.0 |
| `clipr` | 0.8.1 | `progress` | 1.2.3 |
| `clue` | 0.3-68 | `progressr` | 0.19.0 |
| `cluster` | 2.1.8.2 | `promises` | 1.5.0 |
| `clusterProfiler` | 4.18.4 | `proxy` | 0.4-29 |
| `clustree` | 0.5.1 | `ps` | 1.9.3 |
| `codetools` | 0.2-20 | `pscl` | 1.5.9 |
| `colorspace` | 2.1-2 | `purrr` | 1.2.2 |
| `commonmark` | 2.0.0 | `quadprog` | 1.5-8 |
| `ComplexHeatmap` | 2.26.1 | `quantreg` | 6.1 |
| `conflicted` | 1.2.0 | `qvalue` | 2.42.0 |
| `corrplot` | 0.95 | `R.methodsS3` | 1.8.2 |
| `cowplot` | 1.2.0 | `R.oo` | 1.27.1 |
| `cpp11` | 0.5.5 | `R.utils` | 2.13.0 |
| `crayon` | 1.5.3 | `R6` | 2.6.1 |
| `credentials` | 2.0.3 | `ragg` | 1.5.2 |
| `crosstalk` | 1.2.2 | `RANN` | 2.6.2 |
| `curl` | 7.1.0 | `rappdirs` | 0.3.4 |
| `data.table` | 1.18.4 | `rbibutils` | 2.4.1 |
| `DBI` | 1.3.0 | `rcmdcheck` | 1.4.0 |
| `dbplyr` | 2.5.2 | `RColorBrewer` | 1.1-3 |
| `DelayedArray` | 0.36.1 | `Rcpp` | 1.1.1-1.1 |
| `DelayedMatrixStats` | 1.32.0 | `RcppAnnoy` | 0.0.23 |
| `deldir` | 2.0-4 | `RcppArmadillo` | 15.2.7-1 |
| `Deriv` | 4.2.0 | `RcppEigen` | 0.3.4.0.2 |
| `desc` | 1.4.3 | `RcppHNSW` | 0.7.0 |
| `DESeq2` | 1.50.2 | `RcppML` | 0.3.7.1 |
| `devtools` | 2.5.2 | `RcppProgress` | 0.4.2 |
| `diffobj` | 0.3.6 | `RcppRoll` | 0.3.2 |
| `digest` | 0.6.39 | `RcppTOML` | 0.2.3 |
| `dir.expiry` | 1.18.0 | `Rdpack` | 2.6.6 |
| `distributional` | 0.7.0 | `readr` | 2.2.0 |
| `doBy` | 4.7.1 | `readxl` | 1.5.0 |
| `docopt` | 0.7.2 | `reformulas` | 0.4.4 |
| `doParallel` | 1.0.17 | `rematch` | 2.0.0 |
| `DOSE` | 4.4.0 | `rematch2` | 2.1.2 |
| `dotCall64` | 1.2 | `remotes` | 2.5.0 |
| `DoubletFinder` | 2.0.6 | `reprex` | 2.1.1 |
| `downlit` | 0.4.5 | `reshape2` | 1.4.5 |
| `dplyr` | 1.2.1 | `ResidualMatrix` | 1.20.0 |
| `dqrng` | 0.4.1 | `reticulate` | 1.46.0 |
| `dtplyr` | 1.3.3 | `rhdf5` | 2.54.1 |
| `dynamicTreeCut` | 1.63-1 | `rhdf5filters` | 1.22.0 |
| `e1071` | 1.7-17 | `Rhdf5lib` | 1.32.0 |
| `ellipsis` | 0.3.3 | `RhpcBLASctl` | 0.23-42 |
| `enrichplot` | 1.30.5 | `Rhtslib` | 3.6.0 |
| `enrichR` | 3.4 | `rjson` | 0.2.23 |
| `eulerr` | 8.0.0 | `rlang` | 1.2.0 |
| `evaluate` | 1.0.5 | `rmarkdown` | 2.31 |
| `fansi` | 1.0.7 | `ROCR` | 1.0-12 |
| `farver` | 2.1.2 | `roxygen2` | 8.0.0 |
| `fastcluster` | 1.3.0 | `rpart` | 4.1.24 |
| `fastDummies` | 1.7.6 | `rprojroot` | 2.1.1 |
| `fastmap` | 1.2.0 | `rsample` | 1.3.2 |
| `fastmatch` | 1.1-8 | `Rsamtools` | 2.26.0 |
| `fgsea` | 1.36.2 | `RSpectra` | 0.16-2 |
| `fields` | 17.3 | `RSQLite` | 3.53.1 |
| `filelock` | 1.0.3 | `rstatix` | 0.7.3 |
| `fitdistrplus` | 1.2-6 | `rstudioapi` | 0.18.0 |
| `FNN` | 1.1.4.1 | `rsvd` | 1.0.5 |
| `fontawesome` | 0.5.3 | `Rtsne` | 0.17 |
| `fontBitstreamVera` | 0.1.1 | `rversions` | 3.0.0 |
| `fontLiberation` | 0.1.0 | `rvest` | 1.0.5 |
| `fontquiver` | 0.2.1 | `s2` | 1.1.11 |
| `forcats` | 1.0.1 | `S4Arrays` | 1.10.1 |
| `foreach` | 1.5.2 | `S4Vectors` | 0.48.1 |
| `forecast` | 9.0.2 | `S7` | 0.2.2 |
| `foreign` | 0.8-91 | `sass` | 0.4.10 |
| `formatR` | 1.14 | `ScaledMatrix` | 1.18.0 |
| `Formula` | 1.2-5 | `scales` | 1.4.0 |
| `fracdiff` | 1.5-4 | `scater` | 1.38.1 |
| `fs` | 2.1.0 | `scattermore` | 1.2 |
| `furrr` | 0.4.0 | `scatterpie` | 0.2.6 |
| `futile.logger` | 1.4.9 | `sctransform` | 0.4.3 |
| `futile.options` | 1.0.1 | `scuttle` | 1.20.0 |
| `future` | 1.70.0 | `selectr` | 0.5-1 |
| `future.apply` | 1.20.2 | `Seqinfo` | 1.0.0 |
| `gargle` | 1.6.1 | `sessioninfo` | 1.2.3 |
| `gdtools` | 0.5.1 | `Seurat` | 5.5.0 |
| `GeneOverlap` | 1.46.0 | `SeuratDisk` | 0.0.0.9021 |
| `generics` | 0.1.4 | `SeuratObject` | 5.4.0 |
| `GENIE3` | 1.32.0 | `SeuratWrappers` | 0.4.0 |
| `GenomeInfoDb` | 1.46.2 | `sf` | 1.1-1 |
| `GenomicRanges` | 1.62.1 | `shape` | 1.4.6.1 |
| `gert` | 2.3.1 | `shiny` | 1.13.0 |
| `GetoptLong` | 1.1.1 | `SingleCellExperiment` | 1.32.0 |
| `ggbeeswarm` | 0.7.3 | `sitmo` | 2.0.2 |
| `ggdist` | 3.3.3 | `slam` | 0.1-55 |
| `ggforce` | 0.5.0 | `slider` | 0.3.3 |
| `ggfun` | 0.2.0 | `snow` | 0.4-4 |
| `ggiraph` | 0.9.6 | `sourcetools` | 0.1.7-2 |
| `ggnewscale` | 0.5.2 | `sp` | 2.2-1 |
| `ggplot2` | 4.0.3 | `spam` | 2.11-4 |
| `ggplotify` | 0.1.3 | `SparseArray` | 1.10.10 |
| `ggpubr` | 0.6.3 | `SparseM` | 1.84-2 |
| `ggraph` | 2.2.2 | `sparseMatrixStats` | 1.22.0 |
| `ggrastr` | 1.0.2 | `spatial` | 7.3-18 |
| `ggrepel` | 0.9.8 | `spatstat.data` | 3.1-9 |
| `ggridges` | 0.5.7 | `spatstat.explore` | 3.8-1 |
| `ggsci` | 5.0.0 | `spatstat.geom` | 3.8-1 |
| `ggsignif` | 0.6.4 | `spatstat.random` | 3.5-0 |
| `ggtangle` | 0.1.2 | `spatstat.sparse` | 3.2-0 |
| `ggtree` | 4.0.5 | `spatstat.univar` | 3.2-0 |
| `ggvenn` | 0.1.19 | `spatstat.utils` | 3.2-3 |
| `gh` | 1.6.0 | `spData` | 2.3.5 |
| `gitcreds` | 0.1.2 | `spdep` | 1.4-2 |
| `GlobalOptions` | 0.1.4 | `speedglm` | 0.3-4 |
| `globals` | 0.19.1 | `statmod` | 1.5.2 |
| `glue` | 1.8.1 | `stringi` | 1.8.7 |
| `GO.db` | 3.22.0 | `stringr` | 1.6.0 |
| `goftest` | 1.2-3 | `SummarizedExperiment` | 1.40.0 |
| `googledrive` | 2.1.2 | `survival` | 3.8-6 |
| `googlesheets4` | 1.1.2 | `svglite` | 2.2.2 |
| `GOSemSim` | 2.36.0 | `sys` | 3.4.3 |
| `gplots` | 3.3.0 | `systemfonts` | 1.3.2 |
| `graphlayouts` | 1.2.3 | `tensor` | 1.5.1 |
| `gridExtra` | 2.3 | `tester` | 0.3.0 |
| `gridGraphics` | 0.5-1 | `testthat` | 3.3.2 |
| `grr` | 0.9.5 | `textshaping` | 1.0.5 |
| `gson` | 0.1.0 | `tibble` | 3.3.1 |
| `gtable` | 0.3.6 | `tidydr` | 0.0.6 |
| `gtools` | 3.9.5 | `tidygraph` | 1.3.1 |
| `h5mread` | 1.2.1 | `tidyr` | 1.3.2 |
| `harmony` | 2.0.3 | `tidyselect` | 1.2.1 |
| `haven` | 2.5.5 | `tidytree` | 0.4.7 |
| `HDF5Array` | 1.38.0 | `tidyverse` | 2.0.0 |
| `hdf5r` | 1.3.12 | `timechange` | 0.4.0 |
| `hdWGCNA` | 0.4.11 | `timeDate` | 4052.112 |
| `here` | 1.0.2 | `tinytex` | 0.59 |
| `hexbin` | 1.28.5 | `treeio` | 1.34.0 |
| `highr` | 0.12 | `tweenr` | 2.0.3 |
| `Hmisc` | 5.2-5 | `tzdb` | 0.5.0 |
| `hms` | 1.1.4 | `UCell` | 2.14.0 |
| `htmlTable` | 2.5.0 | `UCSC.utils` | 1.6.1 |
| `htmltools` | 0.5.9 | `units` | 1.0-1 |
| `htmlwidgets` | 1.6.4 | `UpSetR` | 1.4.1 |
| `httpuv` | 1.6.17 | `urca` | 1.3-4 |
| `httr` | 1.4.8 | `urlchecker` | 1.0.1 |
| `httr2` | 1.2.2 | `usethis` | 3.2.1 |
| `ica` | 1.0-3 | `utf8` | 1.2.6 |
| `ids` | 1.0.1 | `uuid` | 1.2-2 |
| `igraph` | 2.3.2 | `uwot` | 0.2.4 |
| `impute` | 1.84.0 | `vctrs` | 0.7.3 |
| `ini` | 0.3.1 | `VennDiagram` | 1.8.2 |
| `IRanges` | 2.44.0 | `vipor` | 0.4.7 |
| `irlba` | 2.3.7 | `viridis` | 0.6.5 |
| `isoband` | 0.3.0 | `viridisLite` | 0.4.3 |
| `iterators` | 1.0.14 | `vroom` | 1.7.1 |
| `jquerylib` | 0.1.4 | `waldo` | 0.6.2 |
| `jsonlite` | 2.0.0 | `warp` | 0.2.3 |
| `kableExtra` | 1.4.0 | `WGCNA` | 1.74 |
| `KEGGREST` | 1.50.0 | `whisker` | 0.4.1 |
| `KernSmooth` | 2.23-26 | `withr` | 3.0.2 |
| `knitr` | 1.51 | `wk` | 0.9.5 |
| `labeling` | 0.4.3 | `WriteXLS` | 6.8.0 |
| `lambda.r` | 1.2.4 | `xfun` | 0.58 |
| `later` | 1.4.8 | `xml2` | 1.5.2 |
| `lattice` | 0.22-9 | `xopen` | 1.0.1 |
| `lazyeval` | 0.2.3 | `xtable` | 1.8-8 |
| `leidenbase` | 0.1.37 | `XVector` | 0.50.0 |
| `lifecycle` | 1.0.5 | `yaml` | 2.3.12 |
| `limma` | 3.66.0 | `yulab.utils` | 0.2.4 |
| `listenv` | 0.10.1 | `zellkonverter` | 1.20.1 |
| `littler` | 0.3.23 | `zip` | 2.3.3 |
| `lme4` | 2.0-1 | `zoo` | 1.8-15 |
| `lmtest` | 0.9-40 |  | |
```{=latex}
\normalsize
```


## Python packages (207) {-}


```{=latex}
\footnotesize
```
| Package | Version | Package | Version |
|:--------|:--------|:--------|:--------|
| `adjustText` | 1.3.0 | `notebook` | 6.5.7 |
| `anndata` | 0.12.16 | `notebook_shim` | 0.2.4 |
| `annotated-types` | 0.7.0 | `numba` | 0.65.1 |
| `anyio` | 4.13.0 | `numcodecs` | 0.16.5 |
| `argon2-cffi` | 25.1.0 | `numexpr` | 2.14.1 |
| `argon2-cffi-bindings` | 25.1.0 | `numpy` | 2.4.6 |
| `array-api-compat` | 1.14.0 | `numpy-groupies` | 0.11.3 |
| `arrow` | 1.4.0 | `nvidia-cublas` | 13.1.1.3 |
| `asttokens` | 3.0.1 | `nvidia-cuda-cupti` | 13.0.85 |
| `async-lru` | 2.3.0 | `nvidia-cuda-nvrtc` | 13.0.88 |
| `attrs` | 26.1.0 | `nvidia-cuda-runtime` | 13.0.96 |
| `babel` | 2.18.0 | `nvidia-cudnn-cu13` | 9.20.0.48 |
| `beautifulsoup4` | 4.14.3 | `nvidia-cufft` | 12.0.0.61 |
| `bleach` | 6.3.0 | `nvidia-cufile` | 1.15.1.6 |
| `blosc2` | 4.4.1 | `nvidia-curand` | 10.4.0.35 |
| `cellbender` | 0.3.2 | `nvidia-cusolver` | 12.0.4.66 |
| `certifi` | 2026.5.20 | `nvidia-cusparse` | 12.6.3.3 |
| `cffi` | 2.0.0 | `nvidia-cusparselt-cu13` | 0.8.1 |
| `charset-normalizer` | 3.4.7 | `nvidia-nccl-cu13` | 2.29.7 |
| `click` | 8.4.1 | `nvidia-nvjitlink` | 13.0.88 |
| `comm` | 0.2.3 | `nvidia-nvshmem-cu13` | 3.4.5 |
| `contourpy` | 1.3.3 | `nvidia-nvtx` | 13.0.85 |
| `cuda-bindings` | 13.3.1 | `opt_einsum` | 3.4.0 |
| `cuda-pathfinder` | 1.5.5 | `packaging` | 26.2 |
| `cuda-toolkit` | 13.0.2 | `palantir` | 1.4.4 |
| `cycler` | 0.12.1 | `pandas` | 2.3.3 |
| `debugpy` | 1.8.21 | `pandocfilters` | 1.5.1 |
| `decorator` | 5.3.1 | `parso` | 0.8.7 |
| `defusedxml` | 0.7.1 | `patsy` | 1.0.2 |
| `donfig` | 0.8.1.post1 | `pexpect` | 4.9.0 |
| `elpigraph-python` | 0.3.4 | `pillow` | 12.2.0 |
| `entrypoints` | 0.4 | `pip` | 26.1.2 |
| `executing` | 2.2.1 | `platformdirs` | 4.10.0 |
| `fa2_modified` | 0.4 | `plotly` | 6.7.0 |
| `fast-array-utils` | 1.4.1 | `prometheus_client` | 0.25.0 |
| `fastjsonschema` | 2.21.2 | `prompt_toolkit` | 3.0.52 |
| `filelock` | 3.29.0 | `psutil` | 7.2.2 |
| `fonttools` | 4.63.0 | `ptyprocess` | 0.7.0 |
| `fqdn` | 1.5.1 | `pure_eval` | 0.2.3 |
| `fsspec` | 2026.4.0 | `py-cpuinfo` | 9.0.0 |
| `google-crc32c` | 1.8.0 | `pycparser` | 3.0 |
| `h11` | 0.16.0 | `pydantic` | 2.13.4 |
| `h5py` | 3.16.0 | `pydantic_core` | 2.46.4 |
| `httpcore` | 1.0.9 | `Pygments` | 2.20.0 |
| `httpx` | 0.28.1 | `pynndescent` | 0.6.0 |
| `idna` | 3.18 | `pyparsing` | 3.3.2 |
| `igraph` | 1.0.0 | `pyro-api` | 0.1.2 |
| `ipykernel` | 6.29.5 | `pyro-ppl` | 1.9.1 |
| `ipython` | 9.14.0 | `python-dateutil` | 2.9.0.post0 |
| `ipython-genutils` | 0.2.0 | `python-igraph` | 1.0.0 |
| `ipython_pygments_lexers` | 1.1.1 | `python-json-logger` | 4.1.0 |
| `ipywidgets` | 8.1.8 | `pytz` | 2026.2 |
| `isoduration` | 20.11.0 | `PyYAML` | 6.0.3 |
| `jax` | 0.10.1 | `pyzmq` | 27.1.0 |
| `jaxlib` | 0.10.1 | `referencing` | 0.37.0 |
| `jaxopt` | 0.8.5 | `requests` | 2.34.2 |
| `jedi` | 0.20.0 | `rfc3339-validator` | 0.1.4 |
| `Jinja2` | 3.1.6 | `rfc3986-validator` | 0.1.1 |
| `joblib` | 1.5.3 | `rfc3987-syntax` | 1.1.0 |
| `json5` | 0.14.0 | `rpds-py` | 2026.5.1 |
| `jsonpointer` | 3.1.1 | `rpy2` | 3.6.7 |
| `jsonschema` | 4.26.0 | `rpy2-rinterface` | 3.6.6 |
| `jsonschema-specifications` | 2025.9.1 | `rpy2-robjects` | 3.6.5 |
| `jupyter` | 1.1.1 | `scanpy` | 1.12.1 |
| `jupyter-console` | 6.6.3 | `scFates` | 1.2.4 |
| `jupyter-events` | 0.12.1 | `scikit-learn` | 1.9.0 |
| `jupyter-highlight-selected-word` | 0.2.0 | `scikit-misc` | 0.5.2 |
| `jupyter-lsp` | 2.3.1 | `scipy` | 1.17.1 |
| `jupyter_client` | 7.4.9 | `scverse-misc` | 0.0.7 |
| `jupyter_contrib_core` | 0.4.2 | `seaborn` | 0.13.2 |
| `jupyter_contrib_nbextensions` | 0.7.0 | `Send2Trash` | 2.1.0 |
| `jupyter_core` | 5.9.1 | `session-info2` | 0.4.1 |
| `jupyter_nbextensions_configurator` | 0.6.4 | `setuptools` | 81.0.0 |
| `jupyter_server` | 2.19.0 | `shapely` | 2.1.2 |
| `jupyter_server_terminals` | 0.5.4 | `simpleppt` | 1.1.7 |
| `jupyterlab` | 4.5.7 | `six` | 1.17.0 |
| `jupyterlab_pygments` | 0.3.0 | `soupsieve` | 2.8.4 |
| `jupyterlab_server` | 2.28.0 | `stack-data` | 0.6.3 |
| `jupyterlab_widgets` | 3.0.16 | `statsmodels` | 0.14.6 |
| `kiwisolver` | 1.5.0 | `sympy` | 1.14.0 |
| `lark` | 1.3.1 | `tables` | 3.11.1 |
| `legacy-api-wrap` | 1.5 | `terminado` | 0.18.1 |
| `leidenalg` | 0.12.0 | `texttable` | 1.7.0 |
| `llvmlite` | 0.47.0 | `threadpoolctl` | 3.6.0 |
| `loompy` | 3.0.8 | `tinycss2` | 1.5.1 |
| `lxml` | 6.1.1 | `torch` | 2.12.0 |
| `lxml_html_clean` | 0.4.5 | `tornado` | 6.5.6 |
| `MarkupSafe` | 3.0.3 | `tqdm` | 4.67.3 |
| `matplotlib` | 3.10.9 | `traitlets` | 5.15.0 |
| `matplotlib-inline` | 0.2.2 | `triton` | 3.7.0 |
| `mellon` | 1.7.1 | `typing-inspection` | 0.4.2 |
| `mistune` | 0.8.4 | `typing_extensions` | 4.15.0 |
| `ml_dtypes` | 0.5.4 | `tzdata` | 2026.2 |
| `mpmath` | 1.3.0 | `tzlocal` | 5.4.4 |
| `msgpack` | 1.1.2 | `umap-learn` | 0.5.12 |
| `narwhals` | 2.22.0 | `uri-template` | 1.3.0 |
| `natsort` | 8.4.0 | `urllib3` | 2.7.0 |
| `nbclassic` | 1.3.3 | `wcwidth` | 0.7.0 |
| `nbclient` | 0.10.4 | `webcolors` | 25.10.0 |
| `nbconvert` | 6.5.4 | `webencodings` | 0.5.1 |
| `nbformat` | 5.10.4 | `websocket-client` | 1.9.0 |
| `ndindex` | 1.10.1 | `widgetsnbextension` | 4.0.15 |
| `nest-asyncio` | 1.6.0 | `zarr` | 3.2.1 |
| `networkx` | 3.6.1 |  | |
```{=latex}
\normalsize
```

