---
title: "SingleCell workflow methods"
subtitle: "Stepwise single-cell RNA-seq analysis for wet-lab users"
author: "SingleCell pipeline"
date: "2026-06-22"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    theme: readable
    highlight: tango
---

# Overview

This document describes the current SingleCell workflow as a methods-style
protocol. The pipeline is organized into three chapters:

1. **Chapter 1 - Single-cell processing:** load raw matrices, quality control,
   filtering, doublet removal, integration, clustering, cell-type annotation
   and export of the curated object.
2. **Chapter 2 - Pseudobulk, DESeq2 and networks:** create cell-type subsets,
   assign pseudo-replicates, run DESeq2, render volcano plots, build
   differential expression tables, run GO enrichment, build log2FC heatmaps
   and infer a unified hdWGCNA network from significant genes.
3. **Chapter 3 - Pseudotime:** load the curated AnnData object, select cell
   types, infer a trajectory with Palantir and scFates, plot genes on the
   trajectory and generate gene trend heatmaps.

The workflow is intentionally stepwise. A basic user should mainly edit the
small parameter blocks inside each section, not the helper functions.

# Computational environment

The analysis is designed to run inside the Docker image:

```bash
docker run -it -v ~/projects:/workspace matigara/scrnaseq:latest /bin/bash
```

The workflow scripts assume that the pipeline lives in:

```text
/workspace/workflow_nuevo
```

and that output is written to:

```text
/workspace/resultados
```

The R chapters load:

```{r load-r-libraries}
PIPELINE_DIR <- "/workspace/workflow_nuevo"
DATA_DIR     <- "/workspace/."
base_dir     <- file.path(DATA_DIR, "resultados")

source(file.path(PIPELINE_DIR, "load_libraries.R"))
source(file.path(PIPELINE_DIR, "custom_seurat.R"))
source(file.path(PIPELINE_DIR, "ScRNA_Analysis_Functions.R"))
```

The Python chapter loads:

```{python load-python-libraries}
PIPELINE_DIR = "/workspace/workflow_nuevo"
DATA_DIR     = "/workspace/."
base_dir     = os.path.join(DATA_DIR, "resultados")

exec(open(os.path.join(PIPELINE_DIR, "load_libraries_python.py")).read(), globals())
exec(open(os.path.join(PIPELINE_DIR, "ScRNA_Pseudotime_Functions.py")).read(), globals())
```

# Chapter 1 - Single-cell processing

Chapter 1 produces the curated Seurat object used by the rest of the pipeline.
It is implemented in:

```text
workflow/capitulo1_single_cell.R
```

## Sample definition

The user edits the sample list and colors at the beginning of the chapter.
Each sample points to a raw matrix file and defines the biological condition.

```{r chapter1-samples}
samples <- list(
  list(file = "sample_1/filtered_feature_bc_matrix.h5",
       label = "Sample_1",
       condition = "condition_1"),
  list(file = "sample_2/filtered_feature_bc_matrix.h5",
       label = "Sample_2",
       condition = "condition_2")
)

colors <- c("condition_1" = "#3B82F6",
            "condition_2" = "#EF4444")
```

## Loading and pre-filter QC

The function `load_seurat_samples()` loads all samples and computes quality
metrics such as mitochondrial and chloroplast percentages.

```{r chapter1-load}
mt_pattern <- "^ATMG"
cp_pattern <- "^ATCG"

seurat_list_raw <- load_seurat_samples(
  samples     = samples,
  DATA_DIR    = DATA_DIR,
  USE_CELLBENDER = FALSE,
  mt_pattern  = mt_pattern,
  cp_pattern  = cp_pattern
)
```

## Filtering and doublet detection

The function `filter_seurat_samples()` applies simple QC thresholds and runs
the sample-level filtering logic.

```{r chapter1-filter}
seurat_list <- filter_seurat_samples(
  seurat_list_raw,
  min_features = 200,
  max_mt       = 5
)
```

## Integration, Harmony and clustering

The merged object is normalized, scaled, reduced by PCA and corrected with
Harmony. Clustering is then selected from a user-defined resolution.

```{r chapter1-harmony}
pbmc_harmony <- reduce(seurat_list, merge)
pbmc_harmony <- NormalizeData(pbmc_harmony)
pbmc_harmony <- FindVariableFeatures(pbmc_harmony)
pbmc_harmony <- ScaleData(pbmc_harmony)
pbmc_harmony <- RunPCA(pbmc_harmony)
pbmc_harmony <- RunHarmony(pbmc_harmony, group.by.vars = "orig.ident")
```

```{r chapter1-clustering}
cluster_resolution <- 0.3

pbmc_harmony <- pbmc_harmony |>
  FindNeighbors(reduction = "harmony", dims = 1:30) |>
  FindClusters(resolution = cluster_resolution) |>
  RunUMAP(reduction = "harmony", dims = 1:30)
```

## Marker-based and reference-based annotation

The function `find_markers()` identifies cluster markers. The function
`annotate_by_markers()` applies a manual marker table. When a reference object
is available, `annotate_by_reference()` transfers cell-type labels.

```{r chapter1-annotation}
markers <- find_markers(
  pbmc_harmony,
  output_dir = dir_03
)

pbmc_harmony <- annotate_by_markers(
  pbmc_harmony,
  markers    = markers,
  output_dir = dir_03
)
```

```{r chapter1-reference}
reference_obj <- readRDS(file.path(DATA_DIR, "GSE273033_seuratObj_for_publication.rds"))

pbmc_harmony <- annotate_by_reference(
  query_obj     = pbmc_harmony,
  reference_obj = reference_obj,
  output_dir    = dir_03
)
```

## Optional curation and export

The user can manually group or reassign cell types. The final curated object
is saved as both RDS and H5AD.

```{r chapter1-export}
saveRDS(pbmc_harmony, file.path(dir_objects, "pbmc_harmony_curated.rds"))

export_to_scanpy(
  seurat_obj = pbmc_harmony,
  output_h5ad = file.path(dir_objects, "pbmc_harmony_curated.h5ad")
)
```

# Chapter 2 - Pseudobulk, DESeq2, GO and hdWGCNA

Chapter 2 starts from the curated Seurat object:

```text
workflow/capitulo2_pseudobulk_de.R
```

## Cell-type subsets

The analysis first creates one Seurat object per cell type using
`create_cell_type_subsets()`.

```{r chapter2-subsets}
pseudobulk_annot_col <- "celltype_curated"

cell_type_subsets <- create_cell_type_subsets(
  pbmc_harmony,
  annot_col = pseudobulk_annot_col
)
```

## Pseudo-replicate assignment

Pseudo-replicates are assigned within each condition. The condition column is
used for DESeq2 comparisons; `orig.ident` stores the original sample or library;
`replicate` is created by this step.

```{r chapter2-replicates}
pseudobulk_conditions <- NULL
n_pseudoreps          <- 3

cell_type_subsets_replicates <- assign_pseudoreplicates_batch(
  cell_type_subsets,
  pseudobulk_conditions = pseudobulk_conditions,
  n_pseudoreps          = n_pseudoreps
)
```

The user can check the three relevant metadata columns:

```{r chapter2-replicate-qc}
table(cell_type_subsets_replicates$Pavement_Cell$condition)
table(cell_type_subsets_replicates$Pavement_Cell$replicate)
table(cell_type_subsets_replicates$Pavement_Cell$orig.ident)
```

## Pseudobulk aggregation and DESeq2

The function `run_pseudobulk_deseq2_analysis()` aggregates counts by
pseudo-replicate and runs DESeq2 for each requested comparison.

```{r chapter2-deseq2}
comparisons <- list(
  list(conds = c("condicion1", "condicion2"),
       tag   = "condicion1_vs_condicion2")
)

cell_types_to_analyze <- NULL

deseq2_results <- run_pseudobulk_deseq2_analysis(
  cell_type_subsets_replicates = cell_type_subsets_replicates,
  comparisons                  = comparisons,
  output_dir                   = dir_06,
  cell_types                   = cell_types_to_analyze,
  pseudobulk_dir               = file.path(dir_objects, "pseudobulk_replicas")
)
```

Interpretation:

```text
conds = c("reference", "treatment")
log2FC > 0 means higher expression in treatment
```

## Volcano plots

The function `render_volcano_plots()` uses the DESeq2 tables from a selected
contrast.

```{r chapter2-volcano}
volcano_tag <- "condicion1_vs_condicion2"
padj_cut    <- 0.05
lfc_cut     <- 1

render_volcano_plots(
  results_dir = file.path(dir_06, volcano_tag),
  output_dir  = file.path(dir_06, volcano_tag, "volcano"),
  pdf_name    = paste0("VolcanoPlots_", volcano_tag, ".pdf"),
  padj_cut    = padj_cut,
  lfc_cut     = lfc_cut
)
```

## Differential gene tables

The function `build_differential_tables()` creates the summary tables used by
the heatmap and by the network step.

```{r chapter2-diff-tables}
diff_tables <- build_differential_tables(
  results_dir = file.path(dir_06, volcano_tag),
  output_dir  = file.path(dir_06, volcano_tag),
  prefix      = paste0("diff_table_", volcano_tag),
  padj_cut    = padj_cut,
  lfc_cut     = lfc_cut
)
```

## GO enrichment

GO enrichment is wrapped in `run_go_enrichment_for_contrast()`, so the chapter
stays readable.

```{r chapter2-go}
go_results <- run_go_enrichment_for_contrast(
  results_dir  = file.path(dir_06, volcano_tag),
  output_dir   = file.path(dir_07, volcano_tag),
  orgdb        = org.At.tair.db,
  keytype      = "TAIR",
  go_space     = "BP",
  padj_cutoff  = 0.05
)
```

## Log2FC heatmap

The heatmap compares the differential expression pattern across cell types.

```{r chapter2-heatmap}
heatmap_limits <- c(-5, 5)

build_logfc_heatmap(
  logfc_table  = file.path(dir_06, volcano_tag, "tabla_log2FC_fc1_padj_005.tsv"),
  output_file  = file.path(dir_06, volcano_tag, "log2FC_heatmap.pdf"),
  color_limits = heatmap_limits
)
```

## Unified hdWGCNA on significant genes

The function `run_unified_hdwgcna()` filters the Seurat object to the
significant genes in the DE summary table and runs one unified hdWGCNA network.
This is not one network per cell type; it is one network over the significant
gene subset.

```{r chapter2-hdwgcna}
wgcna_name  <- "unified"
n_metacells <- 25
soft_power  <- NULL

run_unified_hdwgcna(
  seurat_obj    = pbmc_harmony,
  de_table_path = file.path(dir_06, volcano_tag, "tabla_log2FC_fc1_padj_005.tsv"),
  output_dir    = dir_08,
  annot_col     = pseudobulk_annot_col,
  sample_col    = "orig.ident",
  wgcna_name    = wgcna_name,
  n_metacells   = n_metacells,
  soft_power    = soft_power
)
```

## Network export and visualization

The function `plot_hdwgcna_network()` exports edge and node tables and draws the
network. Genes without connections after the TOM threshold are not plotted.

```{r chapter2-network-plot}
tom_threshold <- 0.5
n_hub_label   <- 5

plot_hdwgcna_network(
  hdwgcna_dir   = dir_08,
  output_dir    = dir_08,
  tom_threshold = tom_threshold,
  n_hub_label   = n_hub_label
)
```

Interpretation:

```text
Lower tom_threshold = more edges and more genes.
Higher tom_threshold = stronger edges and fewer genes.
Only connected genes are shown in the final plot.
```

# Chapter 3 - Pseudotime

Chapter 3 is implemented in Python:

```text
workflow/capitulo3_pseudotime.py
```

It uses:

```text
workflow/ScRNA_Pseudotime_Functions.py
```

## Load curated AnnData object

The function `load_curated_object()` reads the H5AD object from Chapter 1,
standardizes coordinate names and saves an overview UMAP.

```{python chapter3-load}
INPUT_H5AD     = os.path.join("/workspace/resultados/objects/pbmc_harmony_curated.h5ad")
ANNOTATION_COL = "celltype_curated"
N_JOBS         = 4

adata, N_JOBS = load_curated_object(
    input_h5ad     = INPUT_H5AD,
    dir_pseudotime = dir_pseudotime,
    annotation_col = ANNOTATION_COL,
    n_jobs         = N_JOBS,
)
```

## Select cell types

The function `preview_trajectory_selection()` subsets the selected cell types
and saves a UMAP preview.

```{python chapter3-selection}
TRAJECTORY_CLUSTERS = ["Pavement Cell", "Stomatal Line"]

adata_sub = preview_trajectory_selection(
    adata          = adata,
    clusters       = TRAJECTORY_CLUSTERS,
    annotation_col = ANNOTATION_COL,
    dir_pseudotime = dir_pseudotime,
)
```

## Infer trajectory

The function `trajectory_run()` defines one parameter set. The function
`run_trajectory_runs()` builds the trajectory using Palantir diffusion maps and
scFates PPT, selects a root cell and assigns pseudotime.

```{python chapter3-trajectory}
ROOT_CLUSTER = "Stomatal Line"

TRAJECTORY_RUNS = [
    trajectory_run(nodes=25, sigma=0.5, lambda_value=30, eigs=3, seed=3),
]

adata_traj, selected_trajectory_dir, trajectory_runs = run_trajectory_runs(
    adata           = adata,
    clusters        = TRAJECTORY_CLUSTERS,
    root_cluster    = ROOT_CLUSTER,
    annotation_col  = ANNOTATION_COL,
    output_base_dir = dir_pseudotime,
    runs            = TRAJECTORY_RUNS,
)
```

## Plot genes on the trajectory

Section 28 is intentionally plain Scanpy code. The user edits only the gene
list.

```{python chapter3-genes}
GENES = ["AT1G22690", "AT4G04890", "AT1G80070"]

gene_plots_dir = os.path.join(selected_trajectory_dir, "gene_plots")
os.makedirs(gene_plots_dir, exist_ok=True)

fig = sc.pl.draw_graph(
    adata_traj,
    color      = GENES,
    title      = GENES,
    show       = False,
    return_fig = True,
)
fig.set_size_inches(18, 18, forward=True)
fig.savefig(os.path.join(gene_plots_dir, "step28_genes_FINAL_18x18.png"),
            dpi=200, facecolor="white")
fig.savefig(os.path.join(gene_plots_dir, "step28_genes_FINAL_18x18.pdf"),
            dpi=200, facecolor="white")
plt.close(fig)
```

## Gene trends

The function `run_step29_gene_trends()` creates two heatmaps:

1. The top `TOP_N` trajectory-associated genes.
2. A customizable highlight list chosen by the user.

```{python chapter3-trends}
TOP_N           = 10
HIGHLIGHT_GENES = ["AT1G22690", "AT4G04890", "AT1G80070"]
ORDERING        = "max"

top_path, highlight_path = run_step29_gene_trends(
    adata        = adata_traj,
    run_dir      = selected_trajectory_dir,
    name         = os.path.basename(selected_trajectory_dir),
    custom_genes = HIGHLIGHT_GENES,
    top_n        = TOP_N,
    ordering     = ORDERING,
    n_jobs       = N_JOBS,
)
```

# Principal helper functions

The workflow keeps complex implementation inside helper functions. The chapter
scripts remain readable and mostly contain parameters.

## R helper functions

```{r r-helper-list}
load_seurat_samples()
filter_seurat_samples()
find_markers()
annotate_by_markers()
annotate_by_reference()
export_to_scanpy()
create_cell_type_subsets()
assign_pseudoreplicates_batch()
run_pseudobulk_deseq2_analysis()
render_volcano_plots()
build_differential_tables()
run_go_enrichment_for_contrast()
build_logfc_heatmap()
run_unified_hdwgcna()
plot_hdwgcna_network()
```

## Python helper functions

```{python python-helper-list}
load_curated_object()
preview_trajectory_selection()
trajectory_run()
build_pseudotime_trajectory()
run_trajectory_runs()
run_step29_gene_trends()
```

# Output structure

The main output folders are:

```text
resultados/
├── 01_qc/
├── 02_clustering/
├── 03_annotation/
├── 04_gene_expression/
├── 05_celltype_curation/
├── 06_pseudobulk_deseq2/
├── 07_go_enrichment/
├── 08_networks/
├── 09_pseudotime/
└── objects/
```

# Reproducibility notes

All plot sizes are standardized to 18 x 18 where applicable. The most important
user-edited parameters are kept directly inside each section. The document
intentionally shows the relevant calls in their native language:

```text
R      = Chapters 1 and 2
Python = Chapter 3
```

This keeps the protocol close to how a wet-lab user actually executes the
pipeline: step by step, changing only visible parameters.
